
# A Brief Introduction to R 


Learning objectives:

1. Learn how to assign a value to an object in R.

2. Learn your way around R (using R Studio).

3. Learn how to do simple arithmetic in R.


## First follow along with this video


<iframe width="560" height="315" src="https://youtube.com/embed/UYclmg1_KLk" frameborder="0" allowfullscreen></iframe>


## Now see if you can do this on your own


Here are some basic things to try in R: 

- Multiplying, dividing, adding numbers in Rstudio console: 

```{r, eval=FALSE}
5*3
5+3
5/3
```

- Assigning a value to an object: (This assigns x the value of 6) 

```{r, eval=FALSE}
x <- 6 
```

- Install a package:

```{r, eval=F}
install.packages("foreign")
```

- Call a package through its library: 
```{r, eval=F}
library(foreign)
```

This CRAN-package, foreign, will come handy when you read data into R next week!


## Take it to the next level

[Follow this link to the free data camp for R](https://www.datacamp.com/courses/free-introduction-to-r/?utm_source=blog_launch_new_intro&utm_medium=blog&utm_campaign=launch_new_intro)

1. You should see that the first lesson there is called "Intro to basics"
2. Click on "View Chapter Details"
3. Complete the following sections of this chapter
    - How it works
    - Arithmetic with R
    - Variable assignment
    - Variable assignment (2)
    - Variable assignment (3)

