
# Visualizing Data in R

Learning Objectives:

1. Learn how to create simple plots in R.

2. Learn how to interpret the results.

## Let's start with a couple of good videos


<iframe width="560" height="315" src="https://www.youtube.com/embed/Hj1pgap4UOY&t=19s" frameborder="0" allowfullscreen></iframe>

<iframe width="560" height="315" src="https://www.youtube.com/embed/FEAS3akVxD8" frameborder="0" allowfullscreen></iframe>


## Basic visulualization

Now go back to the PSenergy data we used in the last chapter.  Follow those instructions to open the data as an object called <code>energyData</code>.


### Making your first plots

We use R to create many different kinds of plots: 


```{r, eval=F}
hist(energyData$Energy, xlab="Television Stories", main="Title")
boxplot(energyData$Energy, ylab="Television Stories", main="Title")
plot(y=energyData$Energy, x=energyData$Approval,ylab="Television Stories",xlab="Presidential Approval")
plot(x=energyData$Energy, ylab="Television Stories",xlab="Month-Year",type="l")
```

### Saving your plot

You can save your plot by re-creating it within a "pdf device."  In essence, R will create the plot as a new file rather than on your screen.

```{r, eval=F}
library(lattice)
getwd()
pdf("stories_over_time.pdf")
densityplot(energyData$Energy, xlab="Television Stories")
dev.off()
```

Do not forget the <code>dev.off</code> or the file will never be rendered.


## Try a bit more on your own.

Complete chapter 1 of the datacamp course on [data visualization in R](https://www.datacamp.com/courses/data-visualization-in-r).
