# Regression example

Let's work work with a dataset that comes from Brandice Canes-Wrone, Brady David, and John Cogan. 2002. “Out of Step, Out of Office: Electoral Accountability and House Members Voting.” *American Political Science Review* 96 (1): 127 - 140. 



The dataset consists of election returns for incumbent members of congress and includes many variables thought to influence elections and election behavior. 

1. [Download the data here](https://raw.githubusercontent.com/jmontgomery/jmontgomery.github.io/master/PS363Extra/incumbents.csv)
2. [The codebook for this example is here.](jmontgomery.github.io/PS363Extra/CodebookIncumbents.pdf)


Once you have the data, you can load it up and take a peek.
```{r, eval=T}
elections.data<-read.csv("https://tinyurl.com/yb4mztc7")
table(elections.data$year)
```

The dataset contains variables related to elections for the U.S. House, including spending patterns of incumbents and challengers. Data are at the electoral district level. Each observation represents a specific contest. Each line of the data includes information about the district, the incumbent, and the challenger (if there is one). Note that some variables are in logarithmic scale. 


-----

1. We want to test the following hypothesis: **As the challenger’s campaign spending increases, the incumbent’s campaign spending increases**. 

$$H_0: \beta=0$$

$$H_0:  \beta \neq 0$$

2. Next it is always a good idea to take a look at the data.  Make a scatterplot for the two variables of interest WITHOUT the least-squares line. As always, we want to have Have clear axes labels. 

```{r, eval=T}
plot(elections.data$chalspend, elections.data$incspend, 
     xlab="Challenger spending (log)", 
     ylab="Incumbent spending (log)", 
     pch=c(19), # points are solid circles
     main="Incumbent vs. Challenger spending",  # Label at the top
     col=rgb(0,0,0,.1)) ## make the points mostly transparent
```

4. Conduct a regression analysis to test the hypothesis 

```{r, eval=T}
RegModel.1 <- lm(incspend~chalspend, data=elections.data)
summary(RegModel.1)
```


5.  So what does this mean?

The prediction equation for the model is:

$$incspend=10.29+0.22*chalspend$$

This means that for every one unit increase in challenger spending (measured in log(dollars)), we expect a 0.22 increase in incumbent spending.  


We can also see that the $t-value$ for tis coefficient is 33.25, and the p-value is very small.  This means we can reject the null hypothesis of no relationship.


6. Let's Go back to the scatterplot and dd the least-squares line. 

```{r, eval=T}
plot(elections.data$chalspend, elections.data$incspend, 
     xlab="Challenger spending (log)", 
     ylab="Incumbent spending (log)", 
     pch=c(19), # points are solid circles
     main="Incumbent vs. Challenger spending",  # Label at the top
     col=rgb(0,0,0,.1)) ## make the points mostly transparent
abline(lm(incspend~chalspend, data=elections.data))
```




