# Multivariate Regression

First, watch this video:

<iframe width="560" height="315" src="https://www.youtube.com/embed/q1RD5ECsSB0" frameborder="0" allowfullscreen></iframe>


## A bit more on multivariate regressions

What if we’re interested in modeling the relationship between two independent variables and a dependent variable? Linear regression can do that through use of matrices and the regression equations. Fortunately, we have R to solve all the matrix algebra for us.

Multivariate Regression can be used to examine the joint relationship between any number of independent variables and a dependent variable. It is important to remember that all of the variables involved have to be continuous or dichotomous. However, also know that linear regression is quite robust to assumption violations. Thus, we can still get workable results if we violate this.

Generally, we can use this prediction equation:
	
$$\hat{y_i} = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \ldots + \beta_k x_{ik}$$

Linear model in R:

```{r, eval=F}
	lm(formula = y ~ x.1 + x.2 + ... + x.k, data = my.data)
```

We can then apply it to the dataset "mtcars", which comes in R. It includes information about various car models in 1974 (boring, right, but stay with me). You can see the variables by typing

```{r, eval=T}
	help("mtcars")
```


We want to test the very simple theory that the miles per gallon a car gets is a function of its weight and its size (displacement).  
$$MPG = \beta_0 + \beta_1 weight + \beta_2 displacement$$

Linear model in R:

```{r, eval=T}
	data("mtcars")
	fit <- lm(mpg ~ wt + disp, data = mtcars)
	summary(fit)
```

Clearly there is a relationship between mileage and weight and, to a lesser degree, the size of the car.  But keep in mind that we are "controlling for" the other variables.  So, a one unit increase in weight **decreases** MPG by 3.35 *even after accounting for the fact that heavier cars tend to be bigger*.

However, the size (dispersion) variable is not significant in this regression.  Does that mean that bigger cars aren't lower in MPG?  Of course not.


```{r, eval=T}
	data("mtcars")
	fit <- lm(mpg ~ disp, data = mtcars)
	summary(fit)
```

Here we see that there is a clear relationship between displacement and MPG.  The relationship only "goes away" when we look at the effect of a one unit change in *displacement controlling for weight*. From this we can understand that:

- A heavier car that is not any bigger will have worse MPG.
- A bigger car that does not get heavier has a more uncertain relationship with MPG. 




## 3D Representation of the Data

Let's take a look at the data.

```{r, eval=t}
pairs(mtcars[,c("mpg", "wt", "disp")])
```

Here we see that all three of these variables seem to be related to eachother.

Let's tryand see this in three dimensions (you may need to install the <code>scatterplot3d</code> package.)


```{r, eval=T}
library(scatterplot3d)
attach(mtcars)
s3d <-scatterplot3d(wt,disp,mpg, pch=16, highlight.3d=TRUE,type="h", main="3D Scatterplot")
fit <- lm(mpg ~ wt+disp)
```


This plot shows the data points for in a three-dimensional space.  But can we now draw the regression "plane" through the points?

```{r, eval=T}
s3d <-scatterplot3d(wt,disp,mpg, pch=16, highlight.3d=TRUE, type="h", main="3D Scatterplot")
fit <- lm(mpg ~ wt+disp) 
s3d$plane3d(fit, col = 'black')
```

Remember, the regression equation is just a representation of he plane.  And the "tilt" of the plane represents "independent" relationship between each covariate and the outcome "controlling for" the other variables.
