# Multivariate regression example: Teaching about evolution

Teaching evolution can be politically controversial, but is scientifically uncontroversial and necessary for understanding biology. How can we determine if politics is skewing the teaching of evolution in schools? Are there individual biases, institutional biases, or a combination (perhaps an interaction) of both?

Effects of qualifications and experience of individual teachers as well as state-level effects on how many hours biology teachers devote to teaching evolution in class. We want to ask ourselves: 

- Does teaching vary by state?
- Does teaching vary by the experience of the teacher?

The data are from the National Survey of High School Biology Teachers and consist of 854 observations of high school biology teachers who were surveyed in the spring of 2007.  The outcome of interest is the number of hours a teacher devotes to human and general evolution in his or her high school biology class (hrsallev).  To see the explanatory variables, see the [codebook](http://jmontgomery.github.io/PS363Extra/evolution_variables.pdf).

Import the evolution.dta data set into R. Go ahead and open the codebook as well.First we need to recode some of the data to tell R where there are missing values.  The female variable codes 1 if the teacher is female, 0 if the teacher is male, and 9 if the sex of the teacher is unknown.  Recode all entries in the female column listed as 9 as NA, and write the code you used to do this below.

```{r, eval=T}
library(foreign)
evolution<-read.dta("http://jmontgomery.github.io/PS363Extra/evolution.dta"
                    , convert.factors=FALSE)
table(evolution$female) ### The 9's should be coded as missing.
evolution$female[evolution$female == 9] <- NA
```


1. Begin  by  creating  a  multivariate  regression  model  using <code>hrsallev</code> as  your  outcome  variable. For  explanatory variables,  use <code>female, phase1, senior_c, notest_p, biocred3, degr3, evolcourse, certified, idscitrans, confident</code>.


```{r, eval=T}
mod.hours<-lm(hrs_allev~phase1+senior_c+notest_p+female
              +biocred3+degr3+evol_course+certified+idsci_trans
              +confident,data=evolution)
summary(mod.hours)
```



1. Interpret the intercept of the model.

> The intercept is 10.2, meaning that theoretically if all the independent variables took on a value of 0 (the teacher had no degree, wasn't certified, was male, had no credibility, etc), there would still be 10.2 hours taught. This is the baseline from which all the other variables come. 

2. According to the model, what is the effect of taking the college-level evolution course.  How does this compare to the effect of holding an additional degree?

> Evol_course's coefficient is 2.57, meaning that those who took a college-level evolution course taught 2.57 more hours of evolution in classes than those who did not take a college-level evolution course. The effect of holding an additional degree is -0.34, meaning those who hold an additional degree usually teach 0.34 fewer hours of evolution than those who do not hold an additional degree. 
