---
title: "Introduction and orientation"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Politcal Science*
date: '2020'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false



---
# Introduction

  - Who are we?
  - Our goals
  - Tips and suggestions
  - Syllabus
  - Getting started in R

  
---

## Introductions

- Who are you?
- What are you going to study?
- What is the last math class you took?
- What was the most exciting/fun thing you did this summer?


---

## Orientation

Class discussion

- What is data science?
- What is team-based learning?
- How does this relate to politics/policy/political science?


---

## Dramatic reading of the syllabus


---


### 1.1 What the .... ?


You can install R from: http://cran.r-project.org/bin/

- Works on all platforms
- Unless you are compiling your own software (e.g., Ubuntu), you will always want the 'binary' distribution of software.
- Helpful videos in the coursebook to get you started.

<br>
<br>

But you should go ahead and also install Rstudio: https://rstudio.com/

- Has great features for building code.
- Additional functionality for making R work nicely with other programs/languages.
- Much mor on this later.

---

### 1.2 The working environment for R/RStudio

- It is possible to do some things using the menus, but we will mostly use command line entry.
- We will also create scripts that can be run all at once, or using shortcut keys from the script editor.
- Let's open up R studio and look around.


> Now make a new script and save it to an appropriate directory.  Add the following, save, and run.



```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center"}
print("Hello world")
```

<br> Highlight the line and hit command-R (in a Mac) to run without menus.


---

### 1.3: Big picture orientation

- Everything in R is object oriented. Data, functions, inputs and outputs will all be data objects.
- R is case sensitive. For that matter, it is also spelling sensitive.
- R is an open-source platform that comes with many basic functions.
- Additional functionality are available in user-contributed packages that can be downloaded, loaded, and used.
- Lots and lots of people use R, write handouts for class, write websites, etc.  Google is your friend (and your enemy).


---

### 1.4: R as a calculator


```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center"}
5+4 # Addition
```

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", fig.cap="Blh"}
6-3  #Subtraction
```

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center"}
34 / 6 # Division
```


```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center"}
5 * 3 # Multiplication
```



---


```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center"}
5^4 # Exponents
``` 

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center"}
625^(1/4) # More exponents
``` 


```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center"}
11%%2 # modular arithmetic (11 mod 2)
``` 


```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center"}
31 %/% 7 # The integer part of a fraction
``` 


---

R comes with a number of constants prestored that you can use

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=FALSE}
6.25 # numbers
pi # And a few others
NA # Missing value
NULL # Nothing.
0/0 # NaN means "Not a number"
1/0 # Inf means infinity
``` 


R follows the order of operations (Please Excuse My Dear Aunt Sally).  What will these evaluate to?

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=FALSE}
2*(3-4)+2
2*(3-4)+2*(4 + 3)^(1/3)
``` 

---

### 1.5: Objects and named storage

What if we wanted to calculate the % of eligible voters who turnout in the 06 elections in Alaska. (Data source: http://www.electproject.org/2006g)


- The `<-` operator asigns a value to that object.  You can also use "=" mostly.
```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
turnout.ak<-238307/496387 # Turnout divided by voting age population
``` 

- The 'show(object)' will be called if you just type the name
```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
turnout.ak 
``` 

---

Named storage also lets you handle each piece separately

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
total.votes.ak <-  238307
voting.age.population.ak <- 496387
turnout.ak <-total.votes.ak/voting.age.population.ak
turnout.ak
``` 



