---
title: "Getting started with objects"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Politcal Science*
date: '2020'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
#   beamer_presentation: default
#   slidy_presentation: default



---

## Orientation for today

We are going to:

1. Talk about assigning object to names in R.
2. Your first objects: functions, vectors, and functions for vectors
3. Learn about your friend the vector


---

## 1: Review: Objects and named storage

What if we wanted to calculate the % of eligible voters who turnout in the 06 elections in Alaska. (Data source: http://www.electproject.org/2006g)


- The `<-` operator asigns a value to that object.  You can also use "=" mostly.
```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
turnout.ak<-238307/496387 # Turnout divided by voting age population
``` 

--

- The 'show(object)' will be called if you just type the name
```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
turnout.ak 
``` 

---

Named storage also lets you handle each piece separately

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
total.votes.ak <-  238307
voting.age.population.ak <- 496387
turnout.ak <-total.votes.ak/voting.age.population.ak
turnout.ak
``` 

--

But Object names cannot start with numbers
```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=FALSE}
06election<-50000 ## Try this
``` 


---

## 2 Objects in the global environment

- Named objects are stored in the "global environment", which means that they can be accessed at any time by any function you might run.
- They are "global" variables (which makes them different from "local" variables).


--


```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}

Bears<-100
ls()

``` 

---

You can also remove objects

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
rm(Bears)
ls()
``` 

--

And you can remove more than one object at a time like this:

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
Bear1<-Bear2<-52
ls()
rm(list=c("Bear1", "Bear2"))
objects()
``` 


---

Combine `rm` and `ls` to clear the global

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
ls()
rm(list=ls())
ls()
```

---

Some things are present in the working environment, but not shown

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
.x<-"Hide me"
print(.x)
ls()
```

--

Anything that starts with a "." will be there, but are hidden


---

## 3 Functions

- Functions are objects that contain a list of instructions
- We "call" a function to execute a command

--


```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
q
```


--


- `q` is a function with three inputs: save, status, runLast


---

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
log
```

- `log` is a function withtwo inputs:
    - the data
    - the base
    
--

- The default base is `exp(1)`

--

- Note that `exp` is itself a function.


---

Try it out!

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
log(2) # ln(2)
log(2, base=10) # log(2) base 10
log(base = 10, x = 2)
```

--

You can call functions inside eachother.  The computer will evaluate the "inside" one first.

```{r, tidy=FALSE, out.height=".55\\textheight", fig.align="center", eval=TRUE}
exp(log(1)) # e^ln(1) = 1
```


---

## 4: Vectors

The c() function is used to collect/concatenate things together into a vector

```{r}
c(0,7,8)
x<-c(0,7,8) # assign this to a named object
```

---

```{r}
x # we can see x
print(x) #Same thing
show(x) #Same thing
```

--

`Print` and `show` do not always produce the same result for more complex objects.


---

An easy way to make a sequence vector is using the `:` operator. 

```{r}
numbers5to20 <- 5:20 # 
numbers5to20
```

--

Note that it always increments by 1

```{r}
1.5:10 
```

--

You can also concatenate any two vectors to make a new vector

```{r}
c(numbers5to20, x)
```

---

## Functions and vectors


Many functions have been "Vectorized" meaning that they work on each element in the vector

```{r}
numbers5to20*2
```

--

```{r}
sqrt(numbers5to20)
```

--

```{r}
numbers5to20^2
```


---


But when you "Interact" two vectors, they will work "elementwise"

```{r}
numbers5to20*numbers5to20
numbers5to20+numbers5to20
```

--

```{r}
numbers5to20+rev(numbers5to20) 
```

---

Other functions operate on on the whole vector

```{r}
sum(numbers5to20)
mean(numbers5to20)
```


---

## Class activity

Helpful functions here include
- `sin`
- `tan`
- `sqrt`
- `sum` 
- `abs` 

---- 

Some math "reminders"
- $\sum_{i=1}^{10} i$ means "sum of numbers from 1 to 10"
- $\forall i \in [1, 20]$ means "for all integers in the range 1 to 20"
    
    

---

## 5: Getting to know your friend the vector

- Enough with the lame coding.  Let's us some data.  
- *Voting age population* and *total votes* cast for the 50 states and DC in 2006

--

```{r}
vap<-voting.age.population<-c(3481823, 496387, 4582842, 2120139,26955438,
                              3617942,2673154,652189,472143,14085749,6915512,
                              995937,1073799,9600372,4732010,2265860,2068253,
                              3213141,3188765,1033632,4242214,4997677,7620982,
                              3908159,2139918,4426278,731365,1321923,1870315,1012033,
                              6598368,1452962,14838076,6752018,494923,8697456,2697855,
                              2850525,9612380,824854,3303593,594599,4636679,
                              17038979,1797941,487900,5841335,4876661,1421717,
                              4257230,392344)    


total.votes<-tv<-c(NA, 238307, 1553032, 780409,8899059,1586105, 
                   1162391,258053, 122356,4884544, 2143845,348988, 
                   458927,3586292, 1719351,1071509, 864083,1370062, 
                   954896,NA, 1809237, 2243835,3852008, 2217552,NA, 
                   2178278, 411061,610499, 586274,418550, 2315643,568597, 
                   4703830,2036451, 220479,4184072, NA,1399650, NA,392882, 
                   1117311,341105, 1868363,NA, 582561, 263025,2398589, 
                   2085074,473014, 2183155, 196217)
```


---

## Accessing elements of a vector or matrix is usually done with the [] operators

We can access the first element of the vector

```{r}
vap[1] 
```

--

Or the 40th!

```{r}
vap[40] 
```

---

We can also extract several elements at a time using vectors.

```{r}
vap[c(3,6,7)] 
numbers5to20[3:7] 
```

--

We can repeat numbers as well

```{r}
total.votes[c(3,3,2,2)] 
```


---

Negative integers *remove* an element and return the rest. Printing everything but the first element.

```{r}
x[-1] 
```

--

Print everything but the 4th through 51st elements

```{r}
vap[-(4:51)] 
```

---


The 0th index returns nothing

```{r}
vap[0] 
vap[c(0:4)]
```

--

Do **not** mix positive and negative indexes
```{r, eval=FALSE}
x[c(-2,3)] 
```


---

## 6: Vector operations and useful vector functions

Basic vector operations

```{r, eval=FALSE}
x
x*3 #scalar multiplication
y<-x-5 #simple addition and multiplicaiton are done "by element"
y
x^3 # ditto with exponents
y^x # but if each are three elements long, it will execute by element
```

---

## WARNING!!! Vector recycling

When vectors have different lengths, the shorter one is extended by 
repeating the vector.

This means two things: 
1. The vector lengths must be multiples of each other
2. This is a *very* easy way to make a bad, bad mistake.

--

----

Dividing by the wrong elements
```{r}
tv[2:3]/vap[2:5] # this is crazy land.  
```

--

----


This is correct
```{r}
tv[2:5]/vap[2:5] 
```

---

For these kinds of reasons, it's nice to be able to access the length of a vector

--

```{r}
length(tv)
length(vap)
length(tv[3:20])
```

---

## Some work on each element seperately

```{r}
sqrt(vap[1:10]) # square root of ever element in the vector
```

--

```{r}
log(vap[1:10]) #  Natural log of each element
```

--

```{r}
abs(c(-1,2,-1)) # absolute value
```

---

## Others will use the whole vector as its input

```{r}
mean(vap) # the mean voting age population
var(vap)  # sample variance of the vap
```

--

```{r}
max(tv, na.rm=T) # maximum -- ignore missing data
min(tv, na.rm=TRUE) #minimum -- notice that T=TRUE
```


---

```{r}
summary(vap) # A summary of the vector 
sort(tv) # creates a sorted vector
```

```{r}
sum(vap) # Adds the vector together
prod(vap[1:4]) # multiplies the vector
```

---

## Some functions evaluate each element

`is.na` returns a vector of `TRUE` or `FALSE` based on whether or not the item is missing (`NA`)

```{r}
is.na(tv) 
```

--

You can reverse this using the `!`, or **not**, symbol.  What's this?

```{r}
!is.na(tv) 
```


---

Class exercise: What's going on here?

```{r}
mean(tv[!is.na(tv)]/vap[!is.na(tv)])
```

--

What would happen if I did this?

```{r}
mean(tv[!is.na(tv)]/vap[!is.na(vap)])
```



---

## Other functions will create vectors as outputs

Repeat the value 1, 5 times
```{r}
rep(1, 5) 
```

--

----

Make the sequence 1 to 21 moving by increments of 2
```{r}
seq(1, 21, by=2) # Make the sequence 1 to 21 moving by increments of 2
```

---

Repeat the pattern 2, 4, ... 20, twice
```{r}
rep(seq(2,20, by=2), 2) 
```

--

----

Repeat 1, 3 times and 4, twice
```{r}
rep(c(1,4), c(3,2)) # repeat 1, 3 times and 4, twice
```

--

----


Repeat each value 3 times
```{r}
rep(c(1,4), each=3) 
```






