---
title: "Basic data structures"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Politcal Science*
date: '2020'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
#   beamer_presentation: default
#   slidy_presentation: default


---


Run this code and change the sign of every odd number in x
```{r, eval=FALSE}
 x <- sample(-100:100, size = 100)
```
<br>

----


--

Subset 'x' so that it contains only numbers between 20 and 93

<br>

----


--

Make a new variable called 'catX' that is has values '0', '1', and '2' based on whether 'x' is in the following ranges: 0=(-100, -30); 1=(-29, 3), 2=(4-100)

<br>

----


--

Make a new variable that is subset of 'x' based on whether 'catX' is 1 OR 2

<br>

----


--

Make a new variable that is subset of 'x' based on whether 'catX' is 1 OR 2 but also only odd numbers.


---


## Orientation for today

**Last time**

1. Talk about object classes
    + Character class
    + Logical class


<br>



--

**Today**

1. Add more complex data structures
    + Matrix
    + Lists
    + Dataframes

<br>

--

**Next class**

1. Flow control
    + if
    + repeat
    + while
    + for


---

## Let's read these in again to get ourselves started

```{r}
vap<-voting.age.population<-c(3481823, 496387, 4582842, 2120139,26955438,
                              3617942,2673154,652189,472143,14085749,6915512,
                              995937,1073799,9600372,4732010,2265860,2068253,
                              3213141,3188765,1033632,4242214,4997677,7620982,
                              3908159,2139918,4426278,731365,1321923,1870315,1012033,
                              6598368,1452962,14838076,6752018,494923,8697456,2697855,
                              2850525,9612380,824854,3303593,594599,4636679,
                              17038979,1797941,487900,5841335,4876661,1421717,
                              4257230,392344)    


total.votes<-tv<-c(NA, 238307, 1553032, 780409,8899059,1586105, 
                   1162391,258053, 122356,4884544, 2143845,348988, 
                   458927,3586292, 1719351,1071509, 864083,1370062, 
                   954896,NA, 1809237, 2243835,3852008, 2217552,NA, 
                   2178278, 411061,610499, 586274,418550, 2315643,568597, 
                   4703830,2036451, 220479,4184072, NA,1399650, NA,392882, 
                   1117311,341105, 1868363,NA, 582561, 263025,2398589, 
                   2085074,473014, 2183155, 196217)
```


---

## Overview for today

- Simplifying things a bit for exposition, there are five basic data structures in R
    + vector
    + matrix
    + array
    + data.frame
    + list



- We have already discussed vectors
- I am leaving arrays to the online content
- We will focus on matrix, data.frame, and list

--

- Most other objects in R are some versions of these


---

## 4.1: The mighty matrix


A matrix is just a collection of vectors

--

You can combine vectors by column using 'cbind'
```{r}
m1<-cbind(vap, tv) 
head(m1)
```

---

Or you can combine by rows using 'rbind'
```{r}
m2<-rbind(vap, tv)
head(m2)
```

---


We can access by 'matrixname[row, column]'
```{r}
m2[1,2] # first row, second column
m1[,1] # the 1st column
```

---


Can get a submatrix

```{r}
m1[1:5,1:2]
m2[1,1:10]
```


---

```{r}
m2[1,1:10]
```

--

```{r}
m2[1:2, 1:10]
```

--

```{r}
m2[, 1:10] # same as previous line since there are only two rows.
```

--

```{r}
class(m2)
```

---

## 3.2: Matrix attributes, column names, and row names

- Objects in R are simply pieces of data that contain specific attributes.
- Classes define what attributes an object *must* have and *can* have.

--

- Matrices always have an attribute that determines the number of rows and columns


```{r}
dim(m1) 
```

- For any object you can look at it's attributes using (helpfully) the 'attributes' function

```{r}
attributes(m1) 
```

---

We see that the matrix also has another attribute


```{r}
dimnames(m1) 
```

- The top refers to the row names (which do not exist)
- The bottom are the column names (that were borrowed from the vectors used to construct the matrix).


---


You can also access these using functions


```{r}
# Another way to do this
colnames(m1)
colnames(m2)
rownames(m1)
rownames(m2)
```

---

And you can reset these using the same functions

```{r}
colnames(m1)<-c("Voting age population", "Total Votes")
colnames(m1)
```


---

You can also do this with the following

```{r}
dimnames(m1)[[2]][1]<-"Pigglywiggly"
head(m1) 
```

- We have re-named the first column to have the name "Pigglywiggly"

--

- We are able to do this because the output of 'dimnames' is a list
- What's a list?


---

## 3.3: Lists

- Many (and maybe most) objects you are going to see in R are lists or versions of lists.
- Lists are a very flexible way to store loosely structure data



--

Let's make a list that contains a *matrix*, a *vector*, and an *integer*.

```{r}
list.a<-list(m1, vap, 3) 
str(list.a)
```


---

This is the advantage of lists.  They can contain basically anything, even other lists

```{r}
vector1<-c(1,2,3)
gospels<-c("matthew","mark","luke", "john")
my.matrix<-matrix(c(1:20), nrow=4)
my.data<-data.frame(cbind(vap, tv))
my.crazy.list<-list(vector1, gospels, my.matrix, TRUE, list.a)
```

---

<small>
```{r, size="tiny"}
str(my.crazy.list)
```
</small>



---

# 3.4: Getting to know your list friend

Lists have attributes but we haven't set them yet

```{r}
attributes(my.crazy.list) 
```

--

This reports the number of major sub-elements in the list.  

```{r}
length(my.crazy.list) 
```


--

This won't work for complicated lists

```{r}
dim(my.crazy.list) 
```


---

We can give names to elements of a list

```{r}
names(my.crazy.list)<-c("OneTwoThree", "Gospels", 
                        "SmallMat", "OneLogical", "AnotherList")
str(my.crazy.list) 
```

Now each part of the list has a name attribute





---


Skip the steps above by doing ...


```{r}
my.crazy.list<-list(OneTwoThree=vector1,
                    Gospels=gospels, 
                    SmallMat=my.matrix, 
                    OneLogical=TRUE, 
                    AnotherList=list.a)
str(my.crazy.list)
names(my.crazy.list)
```


---

# 3.5: Accessing/adding elements of a list

At least four ways of accessing elements of a list

```{r}
my.crazy.list[[1]]
```


```{r}
my.crazy.list$OneTwoThree
```


---

```{r}
my.crazy.list[1]
```


```{r}
my.crazy.list["OneTwoThree"]
```


---

You can add elements in a similarly confusing number of ways


```{r}
my.crazy.list$hocuspocus<-"hocuspocus"
str(my.crazy.list)
```

---

You can also access access/add to/subtract from the sub-elements themselves. We just add what we know about accessing elements of matrices/vectors/etc. to how we access lists.



```{r}
my.crazy.list[[3]][1,] # first row of my.matrix
my.matrix[1,]  #the same

```

--

But lists don't play well with basic commands. Only components of lists.

```{r, eval=FALSE}
my.crazy.list +2 # not so much
```
```{r}
my.crazy.list[[3]] + 2
```


---

## 3.6: data.frames -- the king of the data world

- Dataframes are like a matrix and a list had a baby.
- They are rectangular like a matrix.
- But each column can be of a different class.
- And you can access elements of the data frame like it's a list.


```{r}
turnout<-tv/vap
voting.data <- data.frame(tv, vap, turnout)
head(voting.data)
str(voting.data)
```

---

## 3.7: Accessing and using varaibles

There are at least four ways to access a variable
```{r}
head(voting.data[[1]])
head(voting.data$tv)
```

---

```{r}
head(voting.data["tv"])
head(voting.data[,1])
```




---

## Access the column names

```{r}
names(voting.data)
colnames(voting.data)
```

---

# We can change back and forth between a matrix and a data frame

```{r}
as.data.frame(my.matrix)
data.frame(my.matrix)
```

---
```{r}
head(as.matrix(voting.data))
```

---

- Sometimes all of this can be a bit cumbersome
- The 'with' command will run a function with the dataset slightly easier

```{r}
mean(voting.data$vap)
with(voting.data, mean(vap)) 
```





