---
title: "Flow Control"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Politcal Science*
date: '2020'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
#   beamer_presentation: default
#   slidy_presentation: default


---

# Flow control: Making your R scripts less horrible

- You will find that your scripts start to get **long**.
- Complex tasks especially start turning into complex code
- *TIP* If you find yourself yourself copying and pasting more than 2-3 times --
think about writing a loop or a function instead.

--

- And if you ever spend more than 20 minutes manually reshaping, editing,
 copying/pasting data that is already encoded and on a computer ---
 then somewhere a fairy dies.

--

Some combination of the basic skills in today's lessons can be used
to solve most of these kinds of problems -- although it may take
time work out how.


---


## Orientation for today

**Last time**

1. Add more complex data structures
    + Matrix
    + Lists
    + Dataframes


<br>



--

**Today**

1. Flow control
    + if
    + repeat
    + while
    + for
    
<br>

--

**Next class**

1. Functions 


---

## 5.1: if, else, ifelse

The basic syntax for an `if` call is

```{r, eval=FALSE}
if(condition){
  commands to run
}
```

- The inputs in the parentheses needs can be anything that returns a logical.  
- You can put anything you want in the braces


---

## The simplest examples possible.

```{r}
if(TRUE) {
  print("I got here")
}
```

--

```{r}
if(FALSE){ 
  print("I can't get here")
}
```


---

## else

You can combine this with `else` 

```{r, eval=FALSE}
if(condition) {
  commands to run when condition is TRUE
} else { # notice that these are on the same line
  commands to run when the condition is FALSE
}
```


--


```{r}
x = 10
if(x > 2) {
  print("X is larger than 2")
} else { 
  print("X is 2 or smaller")
}
```

--

Replace 10 with 1 and see what happens.


---

But this set up will not play well with vectors.  It runs, but is confusing



```{r}
if (c(3, 1) > 2){
  print("This won't work")
}

```

---

## ifelse

The `ifelse` command works nicely with vectors, but syntax is different (and often less useful)

```{r, eval=FALSE}
ifelse(condition, return when condition T, return when condition F) 
```

```{r}
x = c(0, 2)
ifelse(x > 1, "yes", "no") 
```


---

## WARNING

If your outputs are vectors `ifelse` will work element-wise


```{r}
yes = c("yes1", "yes2")
no = c("no1", "no2")
ifelse(x > 1, yes, no)
```

---

## Style guide

- Note that the braces are not technically necessary if you have only a one line command 
- BUT you should use them anyway so someone can read your code!

```{r}
x = 3
if (x > 2) y = 2 * x else y = 3 * x
y
```


---

## 4.2) repeat and while


- Repeat just repeats commands in the braces until it sees a `BREAK` command
- If you don't include `BREAK` your computer will be in an infinite loop.
- Save your work before using!
- Or maybe just don't use it

```{r, eval=FALSE}
repeat{ stuff to do until it sees BREAK }
```


---



```{r, fig.height=4, fig.align="center"}
# make a blank plot with the limits set by those vectors
plot(NULL, xlim=c(0, 100), ylim=c(0, 1), xlab="x", ylab="1/x") 
x = 1
repeat {
  y = 1 / x
  x = x + 1
  points(x, y)
  if (x == 100) { 
    break 
  }
}
```

---

A while loop is just a repeat, where the break condition is specified at the top


```{r, eval=FALSE}

while(condition){
  commands that repeat until condition flips to FALSE
}

```

---

```{r, fig.height=4, fig.align="center"}
plot(NULL, xlim=c(0, 100), ylim=c(0, 1), xlab="x", ylab="1/x") 
x = 1
while(x < 100) {
  y = 1 / x
  x = x + 1
  points(x, y)
}
```


---

## 4.3) The mighty for loop

- The `for` command is probably the most common flow control option.
- It has three basic parts:
    + An object name that will be used in the following commands
    + A vector that we will "loop over"
    + Commands that will be executed for each value of the vector


```{r, eval=FALSE}

for (name in vector){
  execute these commands using each value of the vector
}
```

---


```{r}
for (monkey in c("Spider", "Howler", "Rhesus")) {
  ## Each loop does the equivalent of: monkey = "Spider" or monkey = "Howler" or ...
  print(monkey)
}
```

---

Or more commonly

```{r}
for (i in 1:10){
  print(i)
}
```

---


```{r, fig.height=4, fig.align="center"}
## for loops are very useful in many situations
plot(NULL, xlim=c(0, 100), ylim=c(0, 1))
for (i in 1:100) {
  points(i, 1 / i)
}
```

---

## next and break

Sometimes you might not want to execute the commands for every element in the vector
use the `next` command to skip (you can also use the `break`)
```{r}
some.odds = NULL
for (i in 1:200) {
  if (i %% 2 == 0) {
    next
  }
  some.odds = c(some.odds, i)
}
some.odds
```


---

## Style guide again

Technically, you don't have to be so formal. But the indenting and braces are there for the protection of your future self.


```{r}
for (i in 1:10) print(i)
```


---

## Activity 6


