---
title: "Functions"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Politcal Science*
date: '2020'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
#   beamer_presentation: default
#   slidy_presentation: default


---


## Orientation for today

**Last time**



<br>

1. Flow control
    + if
    + repeat
    + while
    + for

--

**Today**

1. Functions 
    
<br>

--

**Next class**

1. Version control/documentation
    + Getting up on github
    + Helping your future self


---


# Functions

- So lets go back and talk about functions again
- We have been using functions this whole time, but we can also make our own.
- This both helps you keep your code organized, and helps you better understand other people's functions

---

## The basics

1. Use the 'function' command and assign it to an object
2. If you need, specify the expected inputs
3. Do some stuff
4. Return something (and only one thing) to the global environment.

```{r, eval=FALSE}
my.function = function(x) {
  do stuff here
  return(output)
}
```



---

## Three ways to write a function: One


```{r}
my.function = function(x) { # take in a value of x
  y = x^2 + 3*x - 2 # conduct operations using the input values
  return(y) # return some value. 
  # In this case we are returning a simple numeric value
}
```

---

## Three ways to write the same function: Two


- You don't have to specify a return
- R will just return whatever is on the last line
```{r}
my.function2 = function(x) {
  x^2 + 3*x - 2 
}
```


---

##  Three ways to write the same function: Three

- And you can do it all in one line
- Not good practice, but sometimes convenient

```{r}
my.function3 = function(x) x^2 + 3*x - 2 
```

---

## Let's test it out

```{r}
my.function(c(1))
my.function2(c(1))
my.function3(c(1))
```

---

## And it's magically vectorized!

```{r}
my.function(c(1:20))
my.function2(c(1:20))
my.function3(c(1:20))
```


---

## DO NUMBER 1.  EVERY TIME.  EXPLICIT IS BETTER!

---

## To recap

Functions are just another object.  We can define the functions using:
1. The word function
2. A pair of round parentheses () which enclose the argument list.  The list may be empty.
3. A single statment, or a sequence of statements enclosed in curly braces { }.


---

## Outputs don't need to be through return ... sort of

Sometimes a function will return a value, but other times it will just execute a command.


```{r}
mad.libs = function(noun, location, proper.noun, activity) {
  print(paste("One day, I was looking all over for my", noun))
  print(paste("I decided I must have left it at the", location))
  print(paste("When I got there, I found", proper.noun, "using my", noun, "to", activity))
}
mad.libs("baseball", "lake house", "Andrew Martin", "eat pudding")
```


---

## Setting defaults


You can set default values for some of your arguments or all of them

```{r}
gaga.equation = function(num.rah=2, num.ah=3, num.ga=2, 
                         num.la=2, num.oo=1) {
  rahs = paste(rep("RAH", num.rah), collapse=", ")
  ahs  = paste(rep("AH", num.ah), collapse=", ")
  gas  = paste(rep("GA", num.ga), collapse=", ")
  oo   = paste(rep("OO", num.oo), collapse=", ")
  las  = paste(rep("LA", num.la), collapse=", ")
  paste(rahs, ",", ahs,  "!  ROMA, ROMAMA!", gas, ",", oo, las)
}

```

---

## Bad Romance


```{r}
gaga.equation()
```

--

```{r}
gaga.equation(num.rah=5)
```

---

## Understanding what does not come out

- Note that rahs is not available outside of the function!  
- This is because of scoping.

```{r, eval=FALSE}
print(rahs)
```

---

## Functions are like Vegas

- As we noted -- local variables are not written into the global environment.
- What happens in the function, stays in the function

---

```{r}
f  = function() {
  x = 1 #local x
  g()
  return(x)
}

g = function() {
  x = 2 # local x
  return(x)
}
f()



x = 3 # global x
f()
g()
x
```

---

## HOWEVER: you can pass values downwards in a function chain


```{r}
f = function() {
  x = 1
  y = g(x)
  return(c(x, y))
}
g = function(x) {
  return(x + 2)
}
a = f()
a
```





