---
title: "ggplot"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Politcal Science*
date: '2020'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
#   beamer_presentation: default
#   slidy_presentation: default


---


## Orientation for today

**Last time**



<br>

1. Lot's of vegetables

--

**Today**

1. Basic data viz
    
<br>

--

**Next class**

1. Data wrangling
    + Recoding data
    + Subsetting
    + Merging
    


---

## Intro to ggplot2

- `ggplot2` is a powerful way to build both simple and complex data visualizations
- Takes care of a lot of the stupid aspects of plot building
- Provides a language for layering visual elements
- Has been extended in dozens of ways to handle all kinds of data
- Integrates easily into the rest of the tidyverse
---

## Running example


Data comes from 538.com
```{r}
library(ggplot2)
primaryPolls<-read.csv('https://jmontgomery.github.io/PDS/Datasets/president_primary_polls_feb2020.csv', stringsAsFactors = F)
primaryPolls$start_date<-as.Date(primaryPolls$start_date, "%m/%d/%Y")
primaryPolls<-primaryPolls[primaryPolls$state=="New Hampshire",]
primaryPolls<-primaryPolls[primaryPolls$candidate_name%in%c("Amy Klobuchar", "Bernard Sanders", "Elizabeth Warren", "Joseph R. Biden Jr.", "Michael Bloomberg", "Pete Buttigieg"),]
```


---

## Our first plot: The 2020

```{r, fig.height=5, fig.align='center'}
ggplot(data=primaryPolls)+
  geom_point(mapping = aes(x=start_date, y=pct))
```


---

## Let's unpack that

```{r, fig.height=5, fig.align='center', eval=FALSE}
ggplot(data=primaryPolls)+
  geom_point(mapping = aes(x=start_date, y=pct))
```

- The first line is always `ggplot`, which sets up the basic object that we will layer onto
- Then we use the `+` to add layers.  In this case `geom_points`
- Any `geom` layer requires a `mapping` argument, which itself comes with an `aes` argument explaining what goes on the x and y coordinats.
- Conveniently, the dataset only needs to be specified once.  


---

### Aesthetics

- The `aes` stands for aesthetics.  
- The nice part is we can easily make this more comlex
- And `ggplot` has pretty good defaults to handle things like color choices, legends, etc.

```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls)+
  geom_point(mapping = aes(x=start_date, y=pct, color= candidate_name))
```


---


- Note that `ggplot` automatically chose a unique color.  
- It can do the same with shapes, point size, and transparency (alpha).
- You can also set aesthetic characteristics manually (as I do with alpha here)
- Note that shapes only works by default with 6 categories

```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls)+
  geom_point(mapping = aes(x=start_date, y=pct, shape= candidate_name, color=candidate_name), alpha=.8)
```


---

## Facets

- Perhaps most helpfully we can parse the data by features using a simple line
- `ggplot` will then arrange everything else so it looks pretty OK


```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls)+
  geom_point(mapping = aes(x=start_date, y=pct)) + 
    facet_wrap(~ candidate_name, nrow=2)
```


---



```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls)+
  geom_point(mapping = aes(x=start_date, y=pct)) + 
    facet_wrap(population ~ candidate_name, nrow=2)
```


---

## More `geom` options

```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls)+
  geom_smooth(mapping = aes(x=start_date, y=pct, color=candidate_name)) + 
    facet_wrap(~ candidate_name, nrow=2)
```

---

```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls)+
  geom_smooth(mapping = aes(x=start_date, y=pct, color=candidate_name)) + 
    geom_point(mapping = aes(x=start_date, y=pct, color=candidate_name), alpha=.4) +
    facet_wrap(~ candidate_name, nrow=2)
```

---

- Maybe we want to do this all on one plot?

```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls)+
  geom_smooth(mapping = aes(x=start_date, y=pct, color=candidate_name, linetype=candidate_name))
```


---

- ANd maybe we don't want a legend?

```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls)+
  geom_smooth(mapping = aes(x=start_date, y=pct, group=candidate_name))
```


---


```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls)+
  geom_smooth(mapping = aes(x=start_date, y=pct, color=candidate_name), show.legend=FALSE)
```



---

- We can combine multiple `geom` objects 

```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls)+
  geom_smooth(mapping = aes(x=start_date, y=pct, color=candidate_name))+
   geom_point(mapping = aes(x=start_date, y=pct, color=candidate_name), alpha=.4)
```

---

- But we can also do this in a way easier to change
```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls, mapping=aes(x=start_date, y=pct, color=candidate_name))+
  geom_smooth()+
   geom_point(alpha=.4)
```

---

- And you can add specifics elments to a sub-layer
```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls, mapping=aes(x=start_date, y=pct, color=candidate_name))+
  geom_smooth()+
   geom_point(aes(size=sample_size), alpha=.4)
```



---

## Class exercise

- Build a plot for the South Carolina primaries.
- You will want to allow for Steyer as well as these candidates
- Start slow and build a more complex set of layers

---

## Lot's more to learn

- A lot of other options you are interested require knowing the right `geom` function

```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls, mapping=aes(x=fte_grade))+
  geom_bar()
```

---

- This is also the same as counting the number of obserervations in each bin 
```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls, mapping=aes(x=fte_grade))+
  stat_count()
```


---

- This is also the same as counting the number of obserervations in each bin 
```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls, mapping=aes(x=candidate_name, y=pct))+
  stat_summary(
    fun.ymin=min,
    fun.ymax=max,
    fun.y=median
  )
```

---

- You can also flip the coordinate system fairly quickly

```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls, mapping=aes(x=candidate_name, y=pct))+
  geom_boxplot()
```

---


```{r, fig.height=5, fig.width=7, fig.align='center', eval=TRUE}
ggplot(data=primaryPolls, mapping=aes(x=candidate_name, y=pct))+
  geom_boxplot() +
  coord_flip()
```


---


## Class exercise (and part of the next problem set)

- Alabama, Arkansas, California, Colorado, Maine, Massachusetts, Minnesota, North Carolina, Oklahoma, Tennessee, Texas, Utah, Vermont, and Virginia will all hold their primraries on March 3
- You have been assigned to create a visulaization of the state of the race for this date.
- You will make a plot to show this.
- In addition to the kinds of issues discussed above
    + Change to the minimial theme
    + Figure out how to change the axis labels and legends beyond the defaults
- Visit https://ggplot2.tidyverse.org/reference/ 









