---
title: "dplyr and tidy"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Politcal Science*
date: '2020'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
#   beamer_presentation: default
#   slidy_presentation: default


---


## Orientation for today

**Last time**



**Before**

1. Basic data viz
    
<br>

--

**Today**

1. Data wrangling
    + Recoding data
    + Subsetting
    + Reshaping
    
    
--

**Next class**
1. Relational data
    

---

## Into the tidyverse

- `dplyr` is a handy package for changing data
- `tidyr` is a handy package for reshaping data
- In combination they offer a powerful way to quickly extract insight from your data



---

## Running example


Data comes from fivethirtyeight.com
```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(readr)
primaryPolls<-read_csv('https://jmontgomery.github.io/PDS/Datasets/president_primary_polls_feb2020.csv')
primaryPolls$start_date<-as.Date(primaryPolls$start_date, "%m/%d/%y")
```


---

### Data object


```{r, eval=TRUE, message=FALSE, warning=FALSE}
class(primaryPolls)
```

- A tibble is basically a data frame 
- They are both friendlier and more structured than traditional data frames


---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
primaryPolls
```


---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
print(primaryPolls, width=Inf)
```


---

- You can access elements basically the same way
- They never allow partial matching (which is evil anyways)
- You can use `as_tibble` or `as.data.frame` to jump back and forth


---

## Back to `dplyr`

- Subset data by rows: `filter`
- Reorder by rows: `arrange`
- Subset data by column: `select`
- Create new variables as a function of other variables: `mutate`
- Collapse values down (or extract statistic): `summarise`
- We can use `group_by` to make changes in the scope

---

## 10.1: `filter`

- `filter` allows you to easily subset data using conditions as we have done before using accessors

```{r, eval=TRUE, message=FALSE, warning=FALSE}
filter(primaryPolls, candidate_name == c("Amy Klobuchar"))

```

---

- Or you can include multiple conditions

```{r, eval=TRUE, message=FALSE, warning=FALSE}
filter(primaryPolls, 
       candidate_name == c("Amy Klobuchar"), 
       state=="New Hampshire")

```


- Everything you have already learned about boolean operators applies here.


---

## 10.2: `arrange`

- `arrange`  does the same thing, but organizes data by rows insead of subsetting by rows.

```{r, eval=TRUE, message=FALSE, warning=FALSE}
arrange(primaryPolls, state, pollster_id)
```


---

- Or in descending order

```{r, eval=TRUE, message=FALSE, warning=FALSE}
arrange(primaryPolls, state, desc(pollster_id))
```

---

## 10.3: `select`

- Select is just a much easier way to subset by column


```{r, eval=TRUE, message=FALSE, warning=FALSE}
select(primaryPolls, state, candidate_name, start_date)
```


---

- It comes with a nice syntax for doing this without just listing
- `var1:var20` will select all columns between these two
- `-var1:var20` will select everything *except* that range
- `starts_with("cand")` will select columns that start with "cand"
- Similar functionality for `ends_with`, `contains`, and matches

---

A fun trick to move your favorite variables to the front, but keep it all:

```{r, eval=TRUE, message=FALSE, warning=FALSE}
select(primaryPolls, state, candidate_name, start_date, everything())
```


---

- Note here that `rename` is the mythical easy way to rename a column, although syntax seems backwards

```{r, eval=TRUE, message=FALSE, warning=FALSE}
basicPolls<-select(primaryPolls, state, candidate_name, start_date, pct)
rename(basicPolls, candidate = candidate_name)
```

---


## 10.4: `mutate`: The glory

- `mutate` allows us to create a new variable that is a function of the others

```{r, eval=TRUE, message=FALSE, warning=FALSE}
mutate(basicPolls, proportion=pct/100)
```

---



```{r, eval=TRUE, message=FALSE, warning=FALSE}
transmute(basicPolls, proportion=pct/100)
```


---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
transmute(primaryPolls, numberRespondents=round((pct/100)*sample_size))
```

---


```{r, eval=TRUE, message=FALSE, warning=FALSE}
transmute(primaryPolls, 
          proportion=pct/100,
          numberRespondents=round(proportion*sample_size)
          )
```

---

- Note that you can use a ton of the basic functions we have already covered like `sum`, `mean`, `sqrt`, etc.
- A useful one is `n()`, which just counts the number of observations
- This includes logical comparisons

```{r, eval=TRUE, message=FALSE, warning=FALSE}
mutate(basicPolls, top_tier=(pct>10)*1)
```

---

## 10.5: `summarise`

- We can easily extract summary statistics


```{r, eval=TRUE, message=FALSE, warning=FALSE}
summarise(basicPolls, average_candidate=mean(pct), count=n())
```

---

- This is more powerful when also using the `group_by` function

```{r, eval=TRUE, message=FALSE, warning=FALSE}
basicPolls_grouped<-group_by(basicPolls, candidate_name)
summarise(basicPolls_grouped, average_candidate=mean(pct), count=n())
```

---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
basicPolls_grouped<-group_by(basicPolls, candidate_name, state)
summarise(basicPolls_grouped, average_candidate=mean(pct), count=n())
```

---

## 10.6 Piping

- The tidyverse includes a nice syntax for combining multiple commands so we don't have to create new objects all of the time.
- The `%?%` syntax allows us to pass on the results of one line to another

```{r, eval=TRUE, message=FALSE, warning=FALSE}
basicPolls %>%
  group_by(candidate_name, state) %>%
  summarise(average_candidate=mean(pct), count=n())
```

- See assigned readings for more useful summary variabels and `ungroup`


---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
basicPolls %>%
  group_by(candidate_name, state) %>%
  summarise(average_candidate=mean(pct), count=n()) %>%
  filter(count>10)
```

---


```{r, eval=TRUE, message=FALSE, warning=FALSE}
primaryPolls %>%
  group_by(candidate_name, state) %>%
  summarise(average_candidate=mean(pct), count=n()) %>%
  filter(count>10) %>%
  mutate(average_prop=average_candidate/100) %>%
  select(average_prop, candidate_name, state, count)
```


---

## Activity

- Filter the data so it includes only polls taken in the last two months
- Select down to only the start_date, pct, state, and candidate
- Create a new variabele that is the proportion of respondents in favor
- Find the median level of support for the top 10 candidates by state limited only to candidate/state combinations with at least 5 polls


---

## 10.7: Tidy data

- In many cases the data is not quite organized the way we want.
- Right now we have each poll as a separate row.  But what if we want each candidate to be a row so we can analyze their trends in polls over time?
- Or what if we get the trend line and want to instead reorganize to look at each poll separately?
- The key commands here are `pivot_wider` and `pivot_longer`
- See Chapter 12 for some additional (but less universally useful functions)


---

## 10.8: Pivots


- In our running example we have one entry for each poll.  How could we combine these to list the result from each unique poll togehter?
- The key here is that the *values* of interest are in the `pct` column and the *groupings* of interest are in the candidates column

```{r, eval=TRUE, message=FALSE, warning=FALSE}
nevadaPrimaries<-primaryPolls %>%
 filter(candidate_name %in% c("Amy Klobuchar", "Bernard Sanders", "Elizabeth Warren", "Joseph R. Biden Jr.", "Michael Bloomberg", "Pete Buttigieg")) %>%
  filter(state=="Nevada") %>%
  select(candidate_name, pct, start_date, sample_size)
nevadaPrimaries
```

---


```{r, eval=TRUE, message=FALSE, warning=FALSE}
wideNevada<-pivot_wider(nevadaPrimaries, names_from = candidate_name, values_from = pct)
wideNevada
```


---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
dim(nevadaPrimaries)
dim(wideNevada)
```

---

- Or we could organize it into a time series ....

```{r, eval=TRUE, message=FALSE, warning=FALSE}
timeNevada<-pivot_wider(nevadaPrimaries, id_cols=candidate_name, 
                        names_from = c(start_date), values_from = pct)
timeNevada
```


--- 


---


- Of course sometimes we want to do the reverse using `pivot_longer`
- In this case, we no longer have a single indicator variable for the values of `start_date`


```{r, eval=TRUE, message=FALSE, warning=FALSE}
timeNevada %>%
  select(candidate_name, `2020-01-08`, `2020-01-06`) %>%
  pivot_longer(c(`2020-01-08`, `2020-01-06`), names_to = "start_date_test", values_to = "pct_test")
```


---

### Activity

- Re-organize the dataset so that there is only one row for each candidate-state dyad
- Feel free to limit this down to only the relevant candidates
- Compare the size of this dataset to our original dataset using the `object_size` command.


