---
title: "More on text"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Politcal Science*
date: '2020'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
#   beamer_presentation: default
#   slidy_presentation: default


---


## Orientation for today

**Last time**



1. Relational data
      
<br>

--

**Today**

1. Advanced text topics (`stringr`)

<br>

--

**Next class**

1. Functional programming
    

---

## Get set up


```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
tweets<-read_csv("C:/Users/dl0ck/OneDrive/Spring2020/Political Data Science/Tweets.csv")
```


---

## Let's start by looking at a single tweet

```{r, eval=TRUE, message=FALSE, warning=FALSE}
aTweet<-tweets[1,]$Text
aTweet
```




---

## Let's go through a number of useful tools here


```{r, eval=TRUE, message=FALSE, warning=FALSE}
str_length(aTweet)
words<-str_split(aTweet, pattern = " ")
words
```

---
```{r, eval=TRUE, message=FALSE, warning=FALSE}
str_c(words)
str_c(unlist(words))
```

---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
str_c(unlist(words), "Added")
```

---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
str_c(unlist(words), "Added", sep=" ")
str_c("Before", unlist(words), "Added", sep=" ")
```



---

### Or you may want to asseble it all into all back into a coherent whole

```{r, eval=TRUE, message=FALSE, warning=FALSE}
str_c(unlist(words), collapse=" ")
```




---

### Things you haven't thought about strings

- Strings in memory are different than how they appear on the screen
- You can use the function `writeLines` to see how a function would atually render to a reader (as opposed to how it is stored in the computer.)
- The big confusion is that special characters have to be "escaped" or else they get in the way (From R4DS)

```{r, eval=TRUE, message=FALSE, warning=FALSE}
x <- c("\"", "\\")
x
writeLines(x)
```


- Others to watch out for are `"\n"` and `"\t"` for `next line` and `tab`
- `na` is a problem, so you might use `str_replace_na`

---

## Covering old ground

- `stringr` replicates a lot of the functions we have already used for characters with (sometimes) slightly different syntax

```{r, eval=TRUE, message=FALSE, warning=FALSE}
colors <- c("red", "yellow", "blue", "green", "magenta", "cyan")
str_sub(colors, 1, 3)
str_sub(colors, -3, -1) #count back from the end
str_sub(colors, 1, 10) # Robust to going out of index
```

---

- This can be combined with other functions as in R


```{r, eval=TRUE, message=FALSE, warning=FALSE}
str_sub(colors, 1, 3)<-str_to_upper(str_sub(colors, 1, 3))
colors
```


--

- Let's undo that

```{r, eval=TRUE, message=FALSE, warning=FALSE}
colors<-str_to_lower(colors)
colors
```

---

## Drill


1. Filter down to all of the tweets from Mayor Lyda Krewson in our data
2. Find the mean number of words in her tweets. Find the unique total.
3. Take all of the words she has used and reduce them down to only their first five letters.  Repeat (2) and compare.


---

## Regular expressions

- Most of the things you are going to want to do with characters are going to involve some sort of pattern matching.
- For this you are going to use something called "regular expressions", which has an entire syntax of its own.

--

- Like many things, this is not native to R but a general approach to pattern matching.
- The potentials for pattern matching are not limitless, but can become extremely complex.
- Probably the most frustrating part for you is that the syntax itself uses `(`, `\`, `.`, `*`, `?`, `{`, and other symbols you might want to search for themselves.
- You are going to want to use escapes.  But because `\` is itself part of the regular expressions syntax you need to double escape `\\`.  So to find `.` you need to search for `\\.`  (And you will need to use `\\\\` to search for `\`)

---


- I cannot cover this all today, but here are some things you can do (and are covered in the book ...) but there are entire books on regular expressions:
    - Find a specific pattern
    - Find a approximate pattern
    - Matching only on the start of a string
    - Match only to the end
    - Match only to a complete string
    - Matches based on white space
    - Matches based on sets 
    - Matches based on digits
    - Matches based on patterns *within* the characters (e.g., letter repetition)

---

## Basic tools

- Lot's of basic tasks have functions already to make it easier.
- But remember in the background, that if you want to do complex queries, subsetting, replacements, etc. --- it is nearly always possible with regex.

--

- But let's back up and focus on
    -`str_detect`: TRUE/FALSE on whether there is a match
    - `str_subset`: Subset down to strings with a match
    - `str_count`: Counts number of matches within a string
    - `str_extract`: Pull out the actual matches from the string
    
- In text aalso covers grouped queries, replacment, `str_locate`, and more.  
- *Read* the book.  

---

## Drill

1. Count the number of times that Krewson tweets included the string 'polic'. 
2. Count the number of times that Krewson tweets the *word* 'police' (See 14.3.2).
3. Count the numer of times her twweets include a link (look at http)
4. Subset to tweets that include the string "Black Lives".  See section 14.5 in the book to see how to do this while ignoring the case.


---

## For next class (and the rest of this one)

1. Using this dataset, come up with your own approach for identifying tweets that mentions:
    + Police, cops, and synonyms
    + Black lives matters protest movements
2. For each mayor in the dataset, what number of tweets match those criteria.
3. Using the `mayors` data from last time, show how these summaary statistics relate (if it relates) to the population size of the city.
4. I do not need a publishable quality project here, but ...
    + I expect matching on more than a couple of words.  Indeed, I expect at least a bit of `regex` work.
    + READ THE BOOK for ideas on how to do this better. 
    + LOOK AT THE DATA for ideas of how to extract better. Subset. View. Pull things out. Iterate.
5. We will compare plots at the start of next class.
6. This is your first graded 'in class' project.



