---
title: "pipes, map"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Politcal Science*
date: '2020'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
#   beamer_presentation: default
#   slidy_presentation: default


---


## Orientation for today

**Last time**



1. Advanced text topics (`stringr`)

      
<br>

--

**Today**


1. More on pipes
2. map

<br>

--

**Next class**

1. Functional programming
    

---

## Get set up

Today we are going to use the 2019 data release from the Voter Study Group

- Download: https://www.voterstudygroup.org/publication/2019-voter-survey-full-data-set
- Codebook: https://www.voterstudygroup.org/publication/2019-voter-survey-full-data-set


```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
VSG<-read_csv("~/Downloads/VOTER_Survey_Jan217_Release1-csv.csv")
```


---

## Start simple: Is there a lane?

We are going to:

- Subset the data to just favorability raitings for Elizabeth Warren and Bernie Sanders
- Recode these variables into "approve" and "not approve"
- Find the proportion of respondents who either approve or disapprove of *both*


---

One strategy is just to make new datsets for each step

```{r, eval=TRUE, message=FALSE, warning=FALSE}
VSG.two <- select(VSG, fav_sanders_2019, fav_warren_2019)
VSG.three<-mutate(VSG.two, fav_sanders_binary = (fav_sanders_2019<=2))
VSG.four<-mutate(VSG.three, fav_sanders_binary= na_if(fav_sanders_2019, fav_sanders_2019>4))
```

ugh

---

Another strategy is just to overwrite the same dataset

```{r, eval=TRUE, message=FALSE, warning=FALSE}
VSG <- select(VSG, fav_sanders_2019, fav_warren_2019)
VSG.<-mutate(VSG, fav_sanders_binary = (fav_sanders_2019<=2)*1)
VSG<-mutate(VSG, fav_sanders_binary= na_if(fav_sanders_2019, fav_sanders_2019>4))
```

Still ugh

--

An even more difficult way would be write a function where the output of each stage are passed onto an outer layer.


---

Or we can use the pipes

```{r, eval=TRUE, message=FALSE, warning=FALSE}
VSG<-read_csv("~/Downloads/VOTER_Survey_Jan217_Release1-csv.csv")
VSG %>% 
  select( fav_sanders_2019, fav_warren_2019) %>%
  mutate(fav_sanders_binary= na_if(fav_sanders_2019, fav_sanders_2019>4)) %>%
  mutate(fav_sanders_binary = (fav_sanders_2019<=2)*1) 
```

---
## How does this work?


- Piping works by creating a new function where object manipulations are passed from one to another
- This means that piping won't work in cases where:
    + The function uses the current environment such as assign, load, or get
    + Functions that use lazy evaluation such as tryCatch, try, or supressMessages
    
    
---

## The limites of pipes

- Pipes are not good in all situations
- If you do more than 4 or 5 pipes, go ahead and make in intermediate object.  Otherwise debugging is a pain.
- If you are not actually working on just one object but instead combinig multiple objects, this is just not worth it.
- Where ther are interdepencies (e.g., complex recoding based on multiple variables)

---
 
---    

## Some other tools

- The `magrittr` package comes with some other tools for using in specific situations
- The `%T>%` pipe works by calling the function to the right, but returning the function to the left


---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(magrittr)
VSG %>% 
  select( fav_sanders_2019, fav_warren_2019) %>%
  mutate(fav_sanders_binary= na_if(fav_sanders_2019, fav_sanders_2019>4)) %>%
  mutate(fav_sanders_binary = (fav_sanders_2019<=2)*1) %T>%
  plot()
```


---

- We may also want to use specific elements from a dataframe using the `%$%` operator

```{r, eval=TRUE, message=FALSE, warning=FALSE}
VSG<- VSG %>% 
  select(fav_sanders_2019, fav_warren_2019) %>%
  mutate(fav_sanders_binary= na_if(fav_sanders_2019, fav_sanders_2019>4)) %>%
  mutate(fav_sanders_binary = (fav_sanders_2019<=2)*1) %>%
  mutate(fav_warrem_binary= na_if(fav_warren_2019, fav_warren_2019>4)) %>%
  mutate(fav_warren_binary = (fav_warren_2019<=2)*1)

```

---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(magrittr)
VSG<- VSG %$% 
  plot(table(fav_sanders_binary, fav_warren_binary))
```

---

## Exercise

- Go back to the original dataset
- To through the same analysis above, to compare support for Sanders/Biden
    + Start by writing the full syntax using overwrites in base R
    + Make a new version using pipes.
    + The target is the same plot we made for Sanders/Warren, but for Sanders/Biden.
    
    
---

## Beyond the loop

- Early in the class, we covered the all powerful for loop.
- But we can do better.
- Let's start by making a dataset we want to loop over

```{r, eval=TRUE, message=FALSE, warning=FALSE}
VSG<-read_csv("~/Downloads/VOTER_Survey_Jan217_Release1-csv.csv")
VSG.fav<-VSG %>% 
  select(starts_with("fav")) 
```

---

So we want to loop over these data and recode all of the columns according to our rules

- Anything above a 4 gets moved to NA
- Then we turn them into a binary based on whether they are 2 and less or not

```{r, eval=TRUE, message=FALSE, warning=FALSE}
output<-as.data.frame(matrix(NA, nrow(VSG.fav), ncol(VSG.fav)))
for(i in seq_along(VSG.fav)){
  output[,i]<-(VSG.fav[,i]<=2)*1
}
```


---

## Can we do better?

- We could do this without making a new object

```{r, eval=TRUE, message=FALSE, warning=FALSE}
for(i in seq_along(VSG.fav)){
  VSG.fav[,i]= na_if(VSG.fav[[i]],VSG.fav[[i]]>4)
  VSG.fav[,i]<-(VSG.fav[,i]<=2)*1
}
head(VSG.fav)
```

---

- We can also loop where we aren't sure how long the final output is
- This can be done by adding each element to the end of a vector as we did in some previous lectures
- This is computationally expensive
- A better solution is to add all of them to a list and then combine them back at the end.
- See Chapter 23.3.3 in the book for more
- Or Chapter 23.3.4 for while loops (which we have already covered)


---

## `map` 

- Looping is such a pervasive task, that there are functions to make it easier
- The `map` family will do this, and you can carefully control the output
  + `map` makes a list
  + `map_lgl` makes a logical vector
  + `map_int` makes an integer
  + `map_dbl` makes a double vector
  + `map_chr` makes a character
  
  
---

## Orientation to maps

- These functions takes in a vector as an input and applies that function
- Let's say that we want to find the mean level of favorability in our dataset


```{r, eval=TRUE, message=FALSE, warning=FALSE}
map(VSG.fav, mean, na.rm=TRUE)
```


---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
map_dbl(VSG.fav, mean, na.rm=TRUE)
```

---

```{r, eval=TRUE, message=FALSE, warning=FALSE}
map_chr(VSG.fav, mean, na.rm=TRUE)
```

---

## walk, purrr, and more

- See Chapter 21.8 for a discussion of `walk`
- Main idea here is when you want to use a function for its side operations (e.g., plot) rather than to manipulate the data
- See also Chapter 21.9.1 for some additional functionality that is occasionally useful:
    + `keep`
    + `discard`
    + `some`
    + `every`
- See Chapter 21.9.2 for even more advanced approaches to extracting data
- Note that I will note require this knowledge for the exam, but this is fair game for the problem set.


---


## Group assignment

- Using the tools from this lecture:
    + Go to the variables that start with `inst`, which represent the level of trust that Americans have in different insitutions.
    + Write a simple set of code to recode these variables to accurately reflect NAs
    + Find the mean level of support for each insitutution
    + Find the median level of support for each institution
    + Make a plot showing trust in institutions ordering institutiosn from least trusted to most trusted among the American public



