---
title: "Webscraping 2: JSON, basic http, and Selenium"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Political Science*
date: '2020'
header-includes:
  - \usepackage{amsmath}
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
#   beamer_presentation: default
#   slidy_presentation: default


---

<style type="text/css">
body, td {
   font-size: 14px;
}
code.r{
  font-size: 14px;
}
pre {
  font-size: 4px;
}
</style>


## Orientation for this component

First half



1. Basic programming
2. Data management/creation

--

Then

1. Some machine learning
2. Special focus on causal inference


--

Now

1. Getting data from the web (basics)
2. **Getting data from the web (more tips)**
3. Getting data from APIs

--

Finally

1. Making html documents
2. R Shiny



---
## The cool side of the web -- programs interfacing with programs

https://www.cnn.com/election/2018/results

- Let's look at this with our inspector.  
- Not so easy, huh?
- But this data has to be coming from somewhere.


---
## Network button for Chrom 

- As pages are rendered, often call to other servers and execute scripts.

- The *network* button is there to help diagnose all of the different connections as websites operate.

- Think of it as a "log" of activity.

--

- But for some websites, we can find where data itself is coming from.

- Let's take a look at the network button as this website is loaded.


---

- We can look at various things here.

- I found the following link:

https://data.cnn.com/ELECTION/2018November6/IN/county/S.json

--

- Boy that looks a lot like the data we want!!


---
## JSON

- A lot of the more complex data-driven websites you want these dats are running JavaScript.
- JSON (JavaScript Object Notation) is a generic format for any data object being passed around.
- It is very similar to unstructured lists in R.
- We'll try downloading this manually first and 


---
## Let's try that with manual download

- I used that url and downloaded the file.  Or you can get it from here.

http://politicaldatascience.com/PDS/Datasets/cnnExample.json

- We are going to use the `rjson` package.



---



```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
library(rjson)
Indiana2018<-fromJSON(file="http://politicaldatascience.com/PDS/Datasets/cnnExample.json")
Indiana2018[[1]]
Indiana2018[[2]]
Indiana2018[[3]] # Race and candidate characterisics
str(Indiana2018[[4]][[1]]) # Results for first county
```

---

```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
str(Indiana2018[[4]][[1]]) # Results for first county
```

---
## So what about the rest?

- Pretty easy to guess where the Ohio results are:

https://data.cnn.com/ELECTION/2018November6/OH/county/S.json

- Can we get this easier than manual download?

--

```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
OH_url<-"https://data.cnn.com/ELECTION/2018November6/OH/county/S.json"
Ohio2018<-fromJSON(file=OH_url)
Ohio2018[[2]]
```


---
## End of Part A



---
## Basic http


- HyperText Transfer Protocol is the basic way that data is passed and accessed across the web.
- GET, POST, PUT, DELETE and HEAD are the main commands.
- We will use the `curl` package (and software) to do this in R.



---
## Example `GET`

- We can see the price history of google here

https://finance.yahoo.com/q/hp?s=GOOG


- We could scrape this, but there is a "download data" button that seems pretty promising.
- Link for that is:

https://query1.finance.yahoo.com/v7/finance/download/GOOG?period1=1555174021&period2=1586796421&interval=1d&events=history


- Can we just get that directly?

---

Using `GET`:

```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
library(RCurl)
google_price<-getForm("https://query1.finance.yahoo.com/v7/finance/download/GOOG", period1="1555174021", period2="1586796421", interval="1d", events="history")
google_price<-read.csv(textConnection(google_price))
head(google_price)
```

Using `POST`:

```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
url <- "http://wwwx.cs.unc.edu/~jbs/aw-wwwp/docs/resources/perl/perl-cgi/programs/cgi_stdin.cgi"
json_data <- postForm(url,
        name = "Bob", 
        checkedbox = "spinich",
        submitButton = "Now!",
        textarea = "Some text to send",
        selectitem = "The item",
        radiobutton = "a",
        style = "POST")
json_data2 <- read.csv(textConnection(json_data))
print(json_data2)
```


---
### End of Part B


---
## Selenium

- The website you want may have forms 
- May be intentionally unfreindly to previous techniques
- May involve a giant backend database you don't/cant get all of it.
- Selenium is software designed to allow you to operate a browser like you are a person.  
- Anything you can access/get via normal browsing behavior is generally accessible.
- I am NOT going to help you get configured, but here is a place to start: https://cran.r-project.org/web/packages/RSelenium/vignettes/basics.html



---
## Jump to R

- This won't really work from slides
- http://politicaldatascience.com/PDS/Lectures/SeleniumExample.R
- You need to install `RSelenium` and Selenium server itself.
- RSelenium is a reference class package: 
http://adv-r.had.co.nz/R5.html
- We are going to use this website as our example: https://www.courts.mo.gov/casenet/base/welcome.do




---

## Some warnings

- What do you call exploiting vulnerabilities in a websites code to gain access to data you are not legally permitted to have?

- What do you all sending a sufficient number of bots at a server requesting small amounts of data until the system cannot continue and freezes/collapses/crashes?

--

- With great power come great responsibility.  Do not get yourself arrested!!


--

- Chapters 8 and 9 cover a lot more topics like certificates, cookies, password/logins, and forms.

---

## Group project  

- Your group should identify a topic for their project
- And find where they want to get the data from
- And write up a one paragraph summary of the idea 
- And send that to me and make an appointment (unless I bounce it)
