---
title: "Basics on APIs"
author: |
  | Jacob M. Montgomery
  | *Washington University in St. Louis*
  | *Department of Political Science*
date: '2020'
header-includes:
  - \usepackage{amsmath}
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
#   beamer_presentation: default
#   slidy_presentation: default


---

<style type="text/css">
body, td {
   font-size: 14px;
}
code.r{
  font-size: 14px;
}
pre {
  font-size: 4px;
}
</style>


## Orientation for this component

Last time

1. Getting data from the web (basics)
2. Getting data from the web (more tips)

--

Now


1. **Getting data from APIs**

--

Next

1. Making html documents/website (Dom)
2. R Shiny


--

But Really

**Work on your group projects**


---
## Getting started

- Application program interfaces (APIs) are a way for people (and programs) to access data in standard format 
- Allows others to build products/software around your data without full access to your data.
- Allows for sharing of data in a controlled way.
    - Only access at a given rate
    - Only access pre-specified fields
- Usually much mor systematic and organized than webscraping
    - But also more powerful (can search their entire dataset)
    - Data comes back with expected structure
    - Documentation!!
- Huge enterprises like OpenSecrets, Twitter, Reddit, Wikipedia, and much more provide these.
- Could be used for making "live" webpages that re-render some kind of data viz. when a page is refreshed to represent changing data. 


---
## Remembering from last time


- We briefly discussed using `http` including commands

- `GET`, `POST`, and `PUT` will give you much of what you need.

- We also learned about the JSON object format, which is similar to a list in R.

--

- This time we are going to use the `httr` and `jsonlite` packages.

- The overall enterprise is to:
    1. Use `http` so pass the right arguments to the right API endpint.
    2. Get a JSON object back.
    3. Optionally use that result to structure a query to another endpoint.
    4. Repeat (3) until you have the data you want.

---
## Some API basics/terminology

- APIs have *endpoints* -- think of these as function that are available.
    - In the documentation (which you should read) it will explain what arguments these endpoints accept.
    - These are (sometiems) passed in via the `http` request.
    - They will then return to you a message (did it work) and an object 
    
----
    
- Most APIs have *rate limits* that control how much data they will return to you.  Try not to wine about people who are giving you free data.
- APIs typically *time out* if your request takes too long to execture, or else the endpint might have a workaround.
- Many will require that you have *authentication* often using a *key*
    - Note that some (e.g., Open Secrets and Twitter) require you to register or even apply

---
## Rough outline for today

- I'll run through a very simple toy example
- We'll do something slightly more complex with the open secrets API.  
    - You'll need to register on your own to follow this example fully.
- The only assignment we will have is I'd like each of you to identify *one* potentially interseting API for the class.  Bonus if you show the result of a query.

---
## Where is the space station?

- Here is a silly little API that tell us at any given time where the International Space Station is located.
- http://open-notify.org/
- If you go to the right url, you can see the answer:
http://api.open-notify.org/iss-now.json

---

- JSON object organized as key: value pairs

----

- "timestamp": 1587069040, 
- "message": "success"
-  "iss_position": 
    * "latitude": "-49.1436"
    * "longitude": "25.2204"

----

- Note that the last elment (iss_position) is itself composes of two elements.


---- 

So we are going to:
1. Send a "GET" request
2. Receive:
    
      a. A response code 
      
      b. Response data


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
library("httr"); library("jsonlite")
iss=GET("http://api.open-notify.org/iss-now.json")
length(iss)
```
---

- Not really the same as just navigating there.
- 200 is the "status response" and indicates success.
- Also includes a lot of details about the call (e.g., time and GET command)

--

- But the actual output we wanted is still there
- And we can pull this out as a "list of lists"

```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
library("httr"); library("jsonlite")
fromJSON(rawToChar(iss$content))
```

---
## End part A

Go play!  See if you can figure out how many people are currently in space.

http://open-notify.org/Open-Notify-API/People-In-Space/

---

## APIs with arguments

- So this one is designed to tell you when the ISS is set to pass over a specific location

http://open-notify.org/Open-Notify-API/ISS-Pass-Times/

- But it requires parameters, just like arguments for a function.
- We need:
    - Latitude 
    - Longitude
- We could try to hack together the url ourselves    
    
http://api.open-notify.org/iss-pass.json?lat=40.7&lon=-74

- But `GET` will do that for us

---


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
iss_pass=GET("http://api.open-notify.org/iss-pass.json", 
             query=list(lat=38.6486009, lon=-90.3128798, n=1))
fromJSON(rawToChar(iss_pass$content))
```

- This is when the ISS will pass over Seigle Hall



---

## Ok that's fun, but kind of dumb

- Let's go look at the Google civic API

https://developers.google.com/civic-information/docs/v2/representatives/representativeInfoByAddress


- And look it has a handy GUI thing

---

### Well ... let's look some more

https://console.developers.google.com/apis/credentials?pli=1


- You are going to need to get your authentication key
- And you you have to click a button in their system to confirm/authorize it.
- But you can borrow mine for now.
- Let's go back and look at this

https://developers.google.com/civic-information/docs/v2/representatives/representativeInfoByAddress



---



```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
myKey="AIzaSyDkCLZcCeC_Uc0d2ItcBBTUqJgk7Pl4ivQ"

SK<-GET("https://www.googleapis.com/civicinfo/v2/representatives",
         query=list(address="6186 McPherson Ave. St. Louis",
                    includeOffices="true",
                    levels="country",
                    roles="legislatorLowerBody", 
                    key="AIzaSyDkCLZcCeC_Uc0d2ItcBBTUqJgk7Pl4ivQ"))
SK
```



---


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
fromJSON(rawToChar(SK$content))
```

---


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
fromJSON(rawToChar(SK$content))$officials
```


---
### End Part B

- See if you can get this to work for a different address.
- If you want to go crazy try and get this endpoint to work instead

https://developers.google.com/civic-information/docs/v2/elections/voterInfoQuery


- use electionID="2000"


---
## API packages

- Smart people have made some R packages for specific APIs.
- So real quick I'll show you how to use the `ropensecretsapi`

---

## A first example


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
ropensecretsapi::SetAPIKey("4b774257ece93adea87064e2933ffbae")
params <- list (id="MO")
MoLegislators <- ropensecretsapi::GetLegislatorsData(params)
MoLegislators
```


---
## Just get the cid


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
MoLegislators$response$legislator[[1]]$`@attributes`[1]
```

---


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
MoLegislators$response$legislator[[1]]$`@attributes`[1]
```

---


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
IDsForMO<-rep(NA, length(MoLegislators$response$legislator))
for(i in 1:length(MoLegislators$response$legislator)){
  IDsForMO[i]<-MoLegislators$response$legislator[[i]]$`@attributes`[1]
}
```

---


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
IDsForMO
```

---


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
getTopDonors<-function(x){
    params<-list(cid=x, cycle="2018")
    ropensecretsapi::GetCandContribData(params)
}
getTopDonors(IDsForMO[1])
```

---


```{r, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
MoTopDonors2018<-plyr::llply(IDsForMO, getTopDonors)
str(MoTopDonors2018)
```


---

### Tasks


Group work:

- Talk to your group about your group project
- By the start of next class please have booked a meeting between me and most of your group.


Individual work

- Find a useful API for the class and post it up online. 
- Bonus if you can figure out how to use it.






