# Getting Started in R

## Learning Objectives 

1. Learn how to assign a value to an object in R.

2. Learn your way around R (using R Studio).

3. Learn how to do simple arithmetic in R.

## Assigning values to objects

You can assign values to objects in two ways as shown below. The assignment arrow is functionally equivalent to the equal sign. When assigning values to objects, R will always take the value on the right side of the assignment operator (<- or =) and store it in the object on the left side of the assignment operator. This means that the two lines of commented code in the following snippet do different things. 

Side note: to comment a single line of code in R, use the hashtag or pound sign at the beginning of the line. If you're working in an R script, the commented portion will turn green.
```{r}
x <- 4
x

y = 6
y

#y = x
#x = y
```

It is best practice when coding to avoid "magic numbers" - i.e. all numbers should be stored in named variables so that if we want to change the value, we only have to do so once. This also removes any ambiguity for someone else reading your code who might wonder what the number represents.
```{r}
total.votes.ak <-  238307
voting.age.population.ak <- 496387
turnout.ak <-total.votes.ak/voting.age.population.ak
turnout.ak
``` 

## R as a calculator


```{r}
5+4 # Addition
6-3 # Subtraction
34 / 6 # Division
5 * 3 # Multiplication
5^4 # Exponents
625^(1/4) # More exponents
11%%2 # modular arithmetic (11 mod 2) is the remainder operator
31 %/% 7 # The integer part of a fraction
``` 

R comes with a number of constants prestored that you can use

```{r, eval=FALSE}
6.25 # numbers
pi # And a few others
NA # Missing value
NULL # Nothing.
0/0 # NaN means "Not a number"
1/0 # Inf means infinity
``` 


R follows the order of operations (Please Excuse My Dear Aunt Sally). 
Side note - if you have multiple exponentiation, they execute right to left.

```{r}
2*(3-4)+2
2*(3-4)+2*(4 + 3)^(1/3)
``` 

## The global environment and how to clean it

Named objects are stored in the "global environment", which means that they can be accessed at any time by any function you might run. They are "global" variables which makes them different from "local" variables (variables that can only be accessed from within a certain scope like within a function).
The commands ls() and rm() are used to show or remove variables from the global environment respectively.

```{r}
a <- 1
b <- 2
ls()
rm(a) #this removes a from the global environment
ls()
c <- 3
rm(list = ls()) #this removes all global variables
```

Some things are present in the working environment, but not shown in the global environment

```{r}
.x<-"Hide me"
print(.x)
ls()
```

Anything that starts with a "." will be accessible in your code, but hidden
