# Data Types in R

## Functions
```{r}
log
log(2) # ln(2)
log(2, base=10) # log(2) base 10
log(base = 10, x = 2)
exp(log(1)) # e^ln(1) = 1
```

## Vectors
The c() function is used to collect/concatenate things together into a vector

```{r}
c(0,7,8)
x<-c(0,7,8) # assign this to a named object
```

An easy way to make a sequence vector is using the `:` operator. 

```{r}
numbers5to20 <- 5:20 # 
numbers5to20
```

Note that it always increments by 1

```{r}
1.5:10 
```

You can also concatenate any two vectors to make a new vector

```{r}
c(numbers5to20, x)
```

## Functions and vectors


Many functions have been "Vectorized" meaning that they work on each element in the vector

```{r}
numbers5to20*2
sqrt(numbers5to20)
numbers5to20^2
```

But when you "Interact" two vectors, they will work "elementwise"

```{r}
numbers5to20*numbers5to20
numbers5to20+numbers5to20
numbers5to20+rev(numbers5to20) 
```

Other functions operate on a functions work more directly

```{r}
sum(numbers5to20)
mean(numbers5to20)
```

Accessing elements of a vector or matrix is usually done with the [] operators

We can access the first or fortieth element of vap.

```{r}
numbers5to20[1] 
numbers5to20[40] 
```

We can also extract several elements at a time using vectors.

```{r}
numbers5to20[c(3,6,7)] 
numbers5to20[3:7]
```


We can repeat numbers as well

```{r}
numbers5to20[c(3,3,2,2)] 
```

## Numerics

## Logicals/Booleans

## Matrices

## Lists

## Dataframes