# Data Types in R

## Functions
```{r}
log
log(2) # ln(2)
log(2, base=10) # log(2) base 10
log(base = 10, x = 2)
exp(log(1)) # e^ln(1) = 1
```

## Vectors
The c() function is used to collect/concatenate things together into a vector

```{r}
c(0,7,8)
x<-c(0,7,8) # assign this to a named object
```

An easy way to make a sequence vector is using the `:` operator. 

```{r}
numbers5to20 <- 5:20 # 
numbers5to20
```

Note that it always increments by 1

```{r}
1.5:10 
```

You can also concatenate any two vectors to make a new vector

```{r}
c(numbers5to20, x)
```

## Functions and vectors


Many functions have been "Vectorized" meaning that they work on each element in the vector

```{r}
numbers5to20*2
sqrt(numbers5to20)
log(numbers5to20)
abs(numbers5to20)
numbers5to20^2
```

But when you "Interact" two vectors, they will work "elementwise"

```{r}
numbers5to20*numbers5to20
numbers5to20+numbers5to20
numbers5to20+rev(numbers5to20) 
data <- c(NA, 4, 7, NA, 19)
is.na(data) #tells us true or false whether each value is NA
!is.na(data) #true if the value is *not* NA
```

Other functions operate on a functions work more directly

```{r}
sum(numbers5to20)
prod(numbers5to20) #multiples all elements together
mean(numbers5to20)
var(numbers5to20)
max(numbers5to20, na.rm=T) # maximum -- ignore missing data
min(numbers5to20, na.rm=TRUE) #minimum -- notice that T=TRUE
summary(numbers5to20)
```

Accessing elements of a vector or matrix is usually done with the [] operators

We can access the first or fortieth element of vap.

```{r}
numbers5to20[1] 
numbers5to20[40] 
```

We can also extract several elements at a time using vectors.

```{r}
numbers5to20[c(3,6,7)] 
numbers5to20[3:7]
```


We can repeat numbers as well

```{r}
numbers5to20[c(3,3,2,2)] 
```
This code prints everything but the 4th through 7th elements by using the negative sign

```{r}
numbers5to20[-(4:7)] 
```

```{r, eval=FALSE}
x
x*3 #scalar multiplication
y<-x-5 #simple addition and multiplicaiton are done "by element"
y
x^3 # ditto with exponents
y^x # but if each are three elements long, it will execute by element
```

Vector recycling

When vectors have different lengths, the shorter one is extended by 
repeating the vector.

This means two things: 
1. The vector lengths must be multiples of each other
2. This is a *very* easy way to make a bad, bad mistake.

You can easily find the length of a vector
```{r}
length(numbers5to20)
```

Other functions will create vectors as outputs


```{r}
rep(1, 5) #Repeat the value 1, 5 times
seq(1, 21, by=2) #Make the sequence 1 to 21 moving by increments of 2
rep(seq(2,20, by=2), 2) #Repeat the pattern 2, 4, ... 20, twice
rep(c(1,4), c(3,2)) #Repeat 1, 3 times and 4, twice
rep(c(1,4), each=3) #Repeat each value 3 times
```

## Numerics

## Logicals/Booleans

## Matrices

## Lists

## Dataframes