# Files, For Loops, and Functions

## The working directory

First we want to change the working directory.  This is the folder where R will save and look for data by default (although you can always override this)

Note that the file path (in Windows at least) requires slashes. You can either use two back slashes to separate folders or one forward slash.

```{r, eval=FALSE}
setwd("~/Dropbox/Classes/R Programming")
## For Windows, will need to be something like
setwd("C:\\Documents and Settings\\Jacob Montgomery\\My Documents\\RWork")
## or
setwd("C:/Documents and Settings/Jacob Montgomery/My Documents/RWork")
## or
setwd(file.path("C:", "Documents and Settings", "Jacob Montgomery", "My Documents", "RWork"))
```

## Input and output

```{r, eval=FALSE}
voting.data<-data.frame(vap, tv)
```

Now let's save our newly created dataset
```{r, eval=FALSE}
dump("voting.data", "voting.data.R") # inputs are string.  Note that you must put the *.R in yourself
```

You might also want to save multiple objects
```{r, eval=FALSE}
dump(c("vap", "tv"), "voting.data.vectors.R")
dump(list=objects(), "everything.R")
## an alternative we can just use the command
save.image("everything.RData") # saves an image of your current workspace
```

Now we can clear our wokspace, and load these objects
```{r, eval=FALSE}
rm(list=ls())
source("voting.data.vectors.R")
ls() # only saved objects present
```

To read/write dataframes from/to a csv (comma separated values) file, the following commands will be useful
```{r, eval=F}
votes.06 <- read.csv("~/Dropbox/Classes/R Programming/R Scripts/VotingData2006.csv", header=T)
write.csv(votes.06, file="VD06.csv")
```
Tips: All of the functions for reading in data are actually using the scan() function. No matter how crazy your data look, you can *always* read it into R by clever use of scan().


## if, else, ifelse

The basic syntax for an `if` call is

```{r, eval=FALSE}
if(condition){
  commands to run
}
```

- The inputs in the parentheses needs can be anything that returns a logical.  
- You can put any code inside the braces

The simplest examples possible:

```{r}
if(TRUE) {
  print("I got here")
}

if(FALSE){ 
  print("I can't get here")
}
```

You can combine this with `else` 

```{r, eval=FALSE}
if(condition) {
  commands to run when condition is TRUE
} else { # notice that these are on the same line
  commands to run when the condition is FALSE
}
```


```{r}
x = 10
if(x > 2) {
  print("X is larger than 2")
} else { 
  print("X is 2 or smaller")
}
```

But this set up will not play well with vectors.  It runs, but is confusing

```{r}
if (c(3, 1) > 2){
  print("This won't work")
}

```

The `ifelse` command works nicely with vectors, but syntax is different. This command is equivalent to a ternary in other languages if you're familiar with that term.

```{r, eval=FALSE}
ifelse(condition, return when condition T, return when condition F) 
```

```{r}
x <- c(0, 2)
ifelse(x > 1, "yes", "no") 
```


WARNING

If your outputs are vectors `ifelse` will work element-wise


```{r}
yes = c("yes1", "yes2")
no = c("no1", "no2")
ifelse(x > 1, yes, no)
ifelse(x < 2, yes, no)
```

- Note that the curly braces are not technically necessary if you have only a one line command 
- BUT you should use them anyway so someone can easily read your code!

```{r}
x <- 3
if (x > 2) y <- 2 * x else y <- 3 * x
y
```

## repeat and while

- Repeat just repeats commands in the braces until it sees a `BREAK` command
- If you don't include `BREAK` your computer will be in an infinite loop.
- Save your work before using!
- Or maybe just don't use it
- `repeat` is equivalent to a `while(true)` loop

```{r, eval=FALSE}
repeat{ stuff to do until it sees BREAK }
```

```{r, fig.height=4, fig.align="center"}
# make a blank plot with the limits set by those vectors
plot(NULL, xlim=c(0, 100), ylim=c(0, 1), xlab="x", ylab="1/x") 
x = 1
repeat {
  y = 1 / x
  x = x + 1
  points(x, y)
  if (x == 100) { 
    break 
  }
}
```


A while loop is just a repeat, where the break condition is specified at the top


```{r, eval=FALSE}
while(condition){
  commands that repeat until condition becomes FALSE
}
```


```{r, fig.height=4, fig.align="center"}
plot(NULL, xlim=c(0, 100), ylim=c(0, 1), xlab="x", ylab="1/x") 
x = 1
while(x < 100) {
  y = 1 / x
  x = x + 1
  points(x, y)
}
```

## for loops

- The `for` command is probably the most common flow control option.
- It has three basic parts:
    + An object name that will be used in the following commands
    + A vector that we will "loop over"
    + Commands that will be executed for each value of the vector


```{r, eval=FALSE}
for (name in vector){
  execute these commands using each value of the vecotr
}
```

```{r}
for (monkey in c("Spider", "Howler", "Rhesus")) {
  ## Each loop does the equivalent of: monkey = "Spider" or monkey = "Howler" or ...
  print(monkey)
}
```

Or more commonly

```{r}
for (i in 1:10){
  print(i)
}
```


```{r, fig.height=4, fig.align="center"}
## for loops are very useful in many situations
plot(NULL, xlim=c(0, 100), ylim=c(0, 1))
for (i in 1:100) {
  points(i, 1 / i)
}
```


### next and break

Sometimes you might not want to execute the commands for every element in the vector
use the `next` command to skip (you can also use the `break`)
```{r}
some.odds = NULL
for (i in 1:200) {
  if (i %% 2 == 0) {
    next
  }
  some.odds = c(some.odds, i)
}
some.odds
```


Technically, you don't have to be so formal. But the indenting and braces are there for the protection of your future self.


```{r}
for (i in 1:10) print(i)
```


## Functions

### The basics

1. Use the 'function' command and assign it to an object
2. If you need, specify the expected inputs
3. Do some stuff
4. (optional) Return something (and only one thing) to the global environment.

```{r, eval=FALSE}
my.function = function(x) {
  do stuff here
  return(output)
}
```

Here's an example:

```{r}
countThrees = function(vector) {
  threes <- 0
  for(i in vector) {
    if(i == 3) {
      threes <- threes + 1
    }
  }
  return(threes)
}

v <- c(1, 2, 3, 4, 3, 3)
countThrees(v)
```

What would happen if we passed in one number instead of a vector?

```{r}
countThrees(2)
countThrees(3)
```

Or a matrix?

```{r}
m <- matrix(1:24, 4, 6)
countThrees(m)
```

If you want to return multiple values in R, you must put them in a vector or list and return the data structure instead.

Remember that a list is essentially a vector that can contain different data types.

Sometimes a function will return a value, but other times it will just execute a command like print or plot.


## Setting defaults

You can set default values for some of your arguments or all of them

```{r}
gaga.equation = function(num.rah=2, num.ah=3, num.ga=2, num.la=2, num.oo=1) {
  rahs = paste(rep("RAH", num.rah), collapse=", ")
  ahs  = paste(rep("AH", num.ah), collapse=", ")
  gas  = paste(rep("GA", num.ga), collapse=", ")
  oo   = paste(rep("OO", num.oo), collapse=", ")
  las  = paste(rep("LA", num.la), collapse=", ")
  paste(rahs, ",", ahs,  "!  ROMA, ROMAMA!", gas, ",", oo, las)
}

gaga.equation()
gaga.equation(num.la=5)
```

If an argument (sometimes called a parameter) does not have a default value, it *must* be specified when the function is called.


## Scope

In the example above, if we tried to print rahs outside of the function, we would get an error. The variable rahs doesn't exist in the global environment.
This concept of limiting which parts of code can access a variable or function is called scope.
```{r, eval=FALSE}
print(rahs)
```

What happens in the function, stays in the function

One exception is that for loop variables can be accessed after the loop.

```{r}
for(i in 1:4) {
  print(i+2)
}
print(i)
```

This for loop could be written as a while loop which would demonstrate why we can access `i` outside the loop.
In fact, when the code is broken down from our nicely readable R code into standard machine code that any computer can run, all for loops
are converted into while loops.

```{r}
i <- 0
while(i < 4) {
  i <- i+1
  print(i+2)
}
print(i)
```
Note that i equals 4 (its last loop value) after the loop is finished.

This is different from functions because the only way to retain access to a local variable that was created in the function is to return it.

```{r}
a <- 10
outer_func <- function() {
  a <- 20
  inner_func <- function() {
    a <- 30
    print(a)
  }
  inner_func()
  print(a)
}
outer_func()
a
```

If you try to run inner\_func(), it will say "Error in inner\_func() : could not find function "inner_func"" because that function is *actually*
a local object inside the function outer\_func. That means that outer\_func is the only place that can use inner\_func.

Scope kind of works like a bunch of one way mirrors. If I write a function, it can access all global variables and local (to the function) variables,
but the global environment can only access global variables. In the above code example, inner\_func can access all of its variables, all of outer\_func's variables,
and the global environment variables as well.

### Advanced scope topics

In R, there's a superassignment operator which is <<-

This assigns the variable to one environment up from its current location. Check out these examples and try to follow along with the environments.

This example will replace the global variable a (which used to store 10) with 20 from the outer\_func code. Printing a at the end of the script will print 20 this time.

```{r}
a <- 10
outer_func <- function() {
  a <<- 20
  inner_func <- function() {
    a <- 30
    print(a)
  }
  inner_func()
  print(a)
}
outer_func()
a
```

This example will replace a from the outer\_func environment with the value 30 so when outer\_func prints a, it will now contain 30.

```{r}
a <- 10
outer_func <- function() {
  a <- 20
  inner_func <- function() {
    a <<- 30
    print(a)
  }
  inner_func()
  print(a)
}
outer_func()
a
```

## Debugging

- `traceback` will help you identify the function that is failing
- `debug`, `debugonce` will go through a function one line at a time
- `browser` This let's you work within the function environment starting at a specified point


Here's an example to help us practice our debugging skills:

```{r}
webData<-url("http://pages.wustl.edu/montgomery/incumbents2.txt")
OOS <- read.table(webData)
summary(OOS)

boxplot(voteshare~inparty, data=OOS[OOS$year==1956,])

plot(voteshare~inparty, data=OOS[OOS$year==1956,])
```

Running a regression by year: Do you see a pattern?


```{r, fig.height=3.5}
output.vector<-NULL
for (i in unique(OOS$year)){
    output.vector[which(unique(OOS$year) == i)]<- 
      lm(voteshare ~ inparty, data=OOS[OOS$year == i,])$coefficients[2]
}
plot(unique(OOS$year), output.vector, type="l", ylab="Coefficient (President's party)", xlab="Year") 
abline(h=0, lty=3)
```

Now we might think -- there is a more general case where I want to get a bunch of seperate regression estimates on defined subsets of the data

```{r, fig.height=3.5, eval=FALSE}
by.var.lm<-function(by.var, formula, data, coef.num){
  output.vector<-NULL
  for (i in unique(by.var)){
    output.vector[which(unique(by.var) == i)] <- 
      lm(formula, data=data[by.var == i,])$coefficients[coef.num]
  }
  return(output.vector)
}
plot(by.var.lm(year, voteshare~inparty, OOS, 2), type="l")
```

This is not working, what to do?

- As a first step, figure out where it's broken using `traceback`
```{r, fig.height=3.5, eval=FALSE}
traceback(by.var.lm(year, voteshare~inparty, OOS, 2))
```

- A second thing to try is to use `debug`
```{r, fig.height=3.5, eval=FALSE}
debug(by.var.lm)
```

- A second thing to try is to put a `broswer` function into the function itself
```{r, fig.height=3.5, eval=FALSE}
browser()
```

- Another thing to try is just put in some various print commands

```{r, fig.height=3.5, eval=FALSE}
by.var.lm<-function(by.var, formula, data, coef.num){
  output.vector<-NULL; print("one")
  for (i in unique(by.var)){ 
    print(i) # I can see where I get to in the loop
    output.vector[which(unique(by.var) == i)]<- 
      lm(formula, data=data[by.var == i,])$coefficients[coef.num]
  }
  print('got out of loop')
  return(output.vector)
}
plot(by.var.lm(year, voteshare~inparty, OOS, 2), type="l")
```

So now I have a function that works, can I use it again?


```{r, fig.height=3.5, eval=FALSE}
website<-url("http://pages.wustl.edu/montgomery/titanic")
titanic<-read.delim(website)
table(titanic$Gender)
table(titanic$Class)
```

Modify the following code to make this work

```{r, eval=FALSE}
by.var.lm(Class, (as.numeric(Survived)-1) ~ Gender, titanic, 2)
```

## Benchmarking and code improvements

```{r}
x <- runif(500)
system.time(sqrt(x))
```

The goal is to see how long a function takes to evaluate

Much better: microbenchmark


```{r}
library(microbenchmark)
microbenchmark(sqrt(x)) # evaluates 100 times per default
microbenchmark(sqrt(x), times=1000)
```

Now we can compare different functions

```{r}
microbenchmark(sqrt(x), 
               x^0.5,
               times=1000)

microbenchmark(sqrt(x), x^0.5, x^(1/2), exp(log(x)/2),
               times=1000)
```

And also completely different functions

```{r}
microbenchmark(sqrt(x), x^4-3*x)
```

For ease of interpretation, if a microbenchmark takes
- 1 millisecond, then 1,000 calls take a second
- 1 microsecond, then 1,000,000 calls take a second
- 1 nanosecond,  then 1,000,000,000 calls take a second

Or use unit=eps for evaluations per second

```{r}
microbenchmark(sqrt(x), x^0.5, x^(1/2), exp(log(x)/2),
               unit="eps")
```

- Evaluating every function takes time
- Evaluating () or {} takes time
- Even specifying useless arguments in functions takes time!

```{r}
f0 <- function() NULL
f1 <- function(a=1) NULL
f2 <- function(a=1, b=2) NULL
f3 <- function(a=1, b=2, c=3) NULL
f4 <- function(a=1, b=2, c=3, d=4) NULL
f5 <- function(a=1, b=2, c=3, d=4, e=5) NULL

microbenchmark(f0(), f1(), f2(), f3(), f4(), f5(), times=10000)
```

Extracting one element of a data frame

```{r}
microbenchmark(
  "[32, 11]" = mtcars[32,11],
  "$carb[32]"	= mtcars$carb[32],
  "[[c(11, 32)]]" = mtcars[[c(11,32)]],
  "[[11]][32]" = mtcars[[11]][32],
  ".subset2" = .subset2(mtcars,11)[32])
```

### Vectorizing

- The key idea behind vectorizing your code is to think about entire
- vectors instead of thinking about their components. Using apply and co
- instead of for loops is a start, but does not really solve this issue.
- Truly vectorized functions will make use of code written in C instead 
 of R. Loops in C are much faster because they have much less overhead.


Addition on each element of a data frame

```{r}
rm(list=ls())
m=5
n=5
matrix1 <- replicate(m, rnorm(n)) # create matrix
matdf <- matdf1 <- data.frame(matrix1) 
matdf

for (i in 1:m) {
  for (j in 1:n) {
    matdf1[i,j] <- matdf1[i,j] + 1.87*cos(.25)*pi # addition
  }
}
matdf1

matdf2<-data.frame(matrix1) 
matdf2 <- matdf2 + 1.87*cos(.25)*pi
matdf2

microbenchmark(
  "loop" = for (i in 1:m) {
    for (j in 1:n) {
      matdf[i,j] <- matdf[i,j] + 1.87*cos(.25)*pi
    }
  }, 
  "vectorized" = matdf <- matdf + 1.87*cos(.25)*pi
)

mat1 <- matrix(abs(rnorm(2500))+pi, ncol=50)
apply(mat1, 1, function(x) sum(x))
rowSums(mat1)

microbenchmark(apply(mat1, 1, function(x) sum(x)),
               rowSums(mat1))
```

- Even for basic tasks, think about the actual calculations you perform

```{r}
mat2 <- matrix(sample(1:7, 90000, replace=T), ncol=300)
mat3 <- matrix(sample(2:6, 90000, replace=T), ncol=300)
ys <- sample(3:5, 300, replace=T)

all.equal(mat2 %*% mat3 %*% ys , mat2 %*% (mat3 %*% ys))
microbenchmark(mat2 %*% mat3 %*% ys,
               mat2 %*% (mat3 %*% ys))
```

- Why?  Think through the dimensionality


### Paste/collapse and copies

```{r}
random_states <- function() {
  paste(sample(state.name,10,replace =TRUE),collapse ="")
}
states10 <- replicate(10, random_states())
states10
states100 <- replicate(100, random_states())

collapse <- function(states) {
  out <- ""
  for (x in states) {
    out <- paste0(out, x) # same as paste(..., sep="", collapse)
  }
  out
}

microbenchmark(
  "loop10" = collapse(states10),
  "vec10" = paste(states10, collapse =""),
  "loop100" = collapse(states100),
  "vec100" = paste(states100, collapse ="")
)
```

PRO TIP:

Allocate memory and fill, don't append to the end

- Here, we are not only getting around using the loop, but also avoiding copies. 
- Whenever you append(), cbind(), rbind(), or paste() to create a bigger object, R must first allocate space for the new object and then copy the old object to its new home. 
- If you're repeating this many times, like in a for loop, this can be quite computationally expensive.
- Instead, allocate an object of the largest size you think you'll need, and fill it up as you go.
- Fun fact: in Java (another programming language), arrays function the same way as vectors in R. Java has a data structure called an ArrayList,
which allows users to add and remove as needed. When it gets full, it doubles the size/capacity of the underlying array and copies everything over. When an element is removed, 
it doesn't decrease the size of the underlying array because copying things over and allocating new space is inefficient.