# More useful packages

## `stringr`

`stringr` is a tidy package that is used for processing strings and text data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
tweets<-read_csv("C:/Users/maria/OneDrive/WUSTL/2019-20/PDS/Tweets.csv")

aTweet<-tweets[1,]$Text
aTweet

str_length(aTweet)
words<-str_split(aTweet, pattern = " ")
words

str_c(words)
str_c(unlist(words))

str_c(unlist(words), "Added")

str_c(unlist(words), "Added", sep=" ")
str_c("Before", unlist(words), "Added", sep=" ")
```

Or you may want to assemble it all into all back into a coherent whole

```{r, eval=TRUE, message=FALSE, warning=FALSE}
str_c(unlist(words), collapse=" ")
```

### Things you haven't thought about strings

- Strings in memory are different than how they appear on the screen
- You can use the function `writeLines` to see how a function would atually render to a reader (as opposed to how it is stored in the computer.)
- The big confusion is that special characters have to be "escaped" or else they get in the way (From R4DS)

```{r, eval=TRUE, message=FALSE, warning=FALSE}
x <- c("\"", "\\")
x
writeLines(x)
```

- Others to watch out for are `"\n"` and `"\t"` for `next line` and `tab`
- `na` is a problem, so you might use `str_replace_na`

- `stringr` replicates a lot of the functions we have already used for characters with (sometimes) slightly different syntax

```{r, eval=TRUE, message=FALSE, warning=FALSE}
colors <- c("red", "yellow", "blue", "green", "magenta", "cyan")
str_sub(colors, 1, 3)
str_sub(colors, -3, -1) #count back from the end
str_sub(colors, 1, 10) # Robust to going out of index

str_sub(colors, 1, 3)<-str_to_upper(str_sub(colors, 1, 3))
colors

colors<-str_to_lower(colors)
colors
```

Example:

1. Filter down to all of the tweets from Mayor Lyda Krewson in our data
2. Find the mean number of words in her tweets. Find the total.
3. Take all of the words she has used and reduce them down to only their first five letters.  Repeat (2) and compare.


Question 2

```{r}
hertweets <- filter(tweets, ScreenName == "lydakrewson")["Text"]
numWords <- hertweets %>% 
    rowwise() %>% 
    transmute(numWords= length(unlist(str_split(Text, " "))))
numWords
mean(numWords$numWords)

length(unique(unlist(str_split(hertweets$Text, " "))))
```

Question 3

```{r}
length(unique(str_sub(unlist(str_split(hertweets$Text, " ")), 1, 5)))
```