# Functions and Packages

## `apply` and its friends

### `apply`

`apply` is a fundamental function in R and this family can be dead useful in many situations

```{r, eval=FALSE}
apply(array, margin, function , ...)
```


1. An array (including potentially a matrix)
2. The margin argument controls how each matrix is analyzed.  Should the function execute on each row (margin = 1) or each column (margin = 2)?
3. The function is what you want done on each row/column/whatever.
4. IMPORTANTLY, the ... refers to any arguments you want to pass to the function


```{r}
mat1 <- matrix(rep(seq(4), 4), ncol = 4)
mat1
```

Let's calculate the sum of each row
```{r}
apply(mat1, 1, sum)
```


Now the sum of each column
```{r}
apply(mat1, 2, sum)
```

Usefully, we are not constrained to the functions that already exist in R.  We can write our own functions.  

```{r}
#using a user defined function
sum.plus.2 <- function(x){
	sum(x) + 2
}
```

Now we can use that function on the rows of our matrix
```{r}
apply(mat1, 1, sum.plus.2)
```

- We can generalize this to add some generic number to our sum.
- And we can also create `anonymous` functions that are never assigned into memory
- What's going on here?


```{r}
apply(mat1, 1, function(x, y) sum(x) + y, y=3)
```

- And here?

```{r}
apply(mat1, 2, function(x, y) sum(x) + y, y=5)
```

### `lapply`


- There are several related functions in R that work the same but with different inputs.

```{r, eval=FALSE}
lapply(list, function, ...)
```

- Many functions in R produce lists and even dataframes are related to lists.

```{r}
mat1.df <- data.frame(mat1)
mat1.df
is.list(mat1.df)
```

- So `lapply` can help you work withs lists and data.frames

```{r}
lapply(mat1.df, sum)
```

- Note that the output of this is a list



- Another useful application of the lapply function is with a "dummy sequence". 
- The list argument is the dummy sequence and it is only used to specify how many iterations we would like to have the function executed. 
- When the lapply functions is used in this way it can replace a for loop very easily, although often the `map` functions discussed last class are better for this.


```{r}
unlist(lapply(1:5, function(i) 5+i ))
```

### `sapply`


- This is a "simplified" version of `lapply`
- The key difference is that it changes hat kind of object is returned depending on what the outcomes look like.
    + If the output is a scalar, the result is a vector
    + If the output is all vectors of the same length, it will return a matrix

```{r, eval=FALSE}
sapply(list, function, ..., simplify)
```

```{r}
sapply(mat1.df, function(x, y) sum(x) + y, y = 5)
```

- This is a vector, not a list

### `tapply`

- This is less intuitive, but can be very useful for recoding tasks, handling data, etc.
- The key here is to understand that the "indices" here are the values of some other object.

```{r, eval=FALSE}
tapply(array, indicies, function, ..., simplify)
```


```{r}
x1 <- runif(16)
x1

cat1 <- rep(1:4, 4)
cat1

cat2 <- c(rep(1, 8), rep(2, 8))
cat2
```


```{r}
mat2.df <- data.frame(x1)
names(mat2.df) <- c("x1")
mat2.df$cat1 <- cat1
mat2.df$cat2 <- cat2
mat2.df
```

  
```{r}
tapply(mat2.df$x1, INDEX = mat2.df$cat1, FUN=mean)
```

And you can do this for combinations of values of variables 

```{r}
tapply(mat2.df$x1, list(mat2.df$cat1, mat2.df$cat2), mean)
```

The first cell here is equivalent to:
```{r}
mean(mat2.df$x1[mat2.df$cat1==1 & mat2.df$cat2==1])
```

### `sweep` 


```{r, eval=FALSE}
sweep(array, margin, stats, function, ...)
```

- This is used if you want, for instance, mean center your variables
- The name is horrible, but the idea is that you can alter each variable as a function of the variable. Easiest to understand by example.

```{r}
a <- matrix(runif(100, 1, 2),20)
a.df <- data.frame(a)
a
```


- Now let's subtract the mean from each column

```{r}
a1 <- sweep(a, 2, colMeans(a), "-")
a1[1:5,]
colMeans(a1) ## column means are all now about zero
```


### `by`

```{r, eval=FALSE}
 by(data, INDICES, FUN, ..., simplify = TRUE)
```

`by` is a wrapper for `tapply` that is supposed to make it easier to use



```{r}
tapply(mat2.df$x1, INDEX = mat2.df$cat1, FUN=mean)
byOut <- by(data=mat2.df$x1, INDICES=mat2.df$cat1, mean)
byOut
class(byOut)
```

### `vapply` And more

Base R comes with a variety of related functions that are a variety on a theme

- `vapply` is mucht the same except
- The argument `FUN.VALUE` provides a template for what the output should look like

```{r, eval=FALSE}
vapply(X, FUN, FUN.VALUE, ..., USE.NAMES = TRUE)
```


```{r}
l <- list(a = 1:10, b = 11:20)
l
```


```{r}
l.fivenum <- vapply(X=l, FUN=fivenum, FUN.VALUE=c(Min.=0, "1st Qu."=0, Median=0, "3rd Qu."=0, Max.=0))
class(l.fivenum)
l.fivenum
```

Other options include:
- `replicate`, which executes the same function multiple times
- `mapply`, which is a multivariate version of `sapply`
- `rapply`, which allows for final handling of outputs
- And more, including:
    + ave
    + colMeans
    + rowSums
    + aggregate
    + eapply


## `plyr`

The `plyr` package is designed to make all of this a bit easier to handle by adding the following features

- Consistent naming protocols for functions in terms of what is going in and coming out
- Easy to make parallel
- Built-in error recovery
- Better handling of labels
- Flesible handling of all basic data types
- This content taken from: https://www.r-bloggers.com/a-fast-intro-to-plyr-for-r/

```{r}
library(plyr)
```


Let's make some example data


```{r}
dd<-data.frame(matrix(rnorm(216),72,3),c(rep("A",24),rep("B",24),rep("C",24)),c(rep("J",36),rep("K",36)))
colnames(dd) <- c("v1", "v2", "v3", "dim1", "dim2")
head(dd)
dd
```


The main functions we want to use are a_ply, aaply, adply, alply, d_ply, daply, ddply, dlply, l_ply, laply, llply, m_ply, maply, mdply, mlply


- The first letter for each tells us what kind of input we have
    + a=array
    + d = data.frame
    + l=list
    + m=matrix
- The second letter tells us what we want in terms of output
    + a=array
    + d=data.frame
    + l=list
    + m=matrix
    + _=discard the results
    

Example: 

```{r}
ddply(.data=dd, .variables=c("dim1","dim2"), .fun=function(df) mean(df$v1))
```

```{r}
daply(.data=dd, .variables=c("dim1","dim2"), .fun=function(df)mean(df$v1))
```

Most combinations we want are here

```{r}
l_ply(1:100, identity)
llply(1:100, identity)
```


```{r}
laply(1:100, identity)
ldply(1:100, identity)
```
```{r, echo=FALSE}
detach("package:plyr", unload=T)
```