# Using SQL syntax on an R dataframe

## SQL

SQL (Structured Query Language) is the universally most widely used querying language for databases. There are variations of SQL, 
the most popular being MySQL and PostgreSQL. The syntax is very similar for all versions of SQL.

[Here](https://www.sqltutorial.org/sql-syntax/)'s a helpful tutorial for SQl commands, but a few basic examples of each keyword are shown below:


### `SELECT` and `FROM`

```
SELECT col1, col2, col3 FROM table1

SELECT * FROM table1
```

The first example selects the three columns specified from a table called table1. It will select all rows from table1 and return them in a new, temporary table
unless the result is stored or used somewhere examples

The second example selects all columns (and rows) from table1.

### `WHERE`

```
SELECT movie_title, user, rating 
    FROM movies 
    WHERE rating > 4 AND user NOT NULL
```

This will select all of the rating data (movie_title, user, and rating) from the movies table with a rating > 4 and a not-null user

### Functions


We can use aggregate functions like sum, min, max, count, or average with group by to apply that function to each group.

Remember that your group by variable must be included in your select statement!
```
-- valid
SELECT movie_title, AVG(rating) 
    FROM movies 
    GROUP BY movie_title 

-- invalid (we wouldn't know which avg goes with which movie!!)
SELECT AVG(rating) 
    FROM movies 
    GROUP BY movie_title 
```

We can also specify what we want the resulting aggregate column to be called. The above example would give us nicer output with this code:

```
SELECT movie_title, AVG(rating) AS rating_avg
    FROM movies 
    GROUP BY movie_title 
```


We can limit the number of results returned, find only unique values, and sort our results as well.

```
SELECT DISTINCT student_id, student_full_name 
    FROM assignments 
    ORDER BY student_full_name DESC 
    LIMIT 100
```

This query will grab the first 100 distinct students (uniqueness determined by id and full name) sorted in reverse alphabetical order.


### `HAVING`

`HAVING` is the equivalent function of `WHERE` for aggregate conditions. For example,

```
SELECT movie_title, AVG(rating) AS rating_avg 
    FROM movies 
    GROUP BY movie_title 
    HAVING rating_avg < 2
```

This pulls all the movies and their average ratings that had an average rating less than 2.

### `WITH`/`AS` and Subqueries

Subqueries are very powerful and come in two syntatical forms. They allow us to create temporary tables which only exist for that query 
then disappear.

```
SELECT employee_id, first_name, last_name, salary 
    FROM employees 
    WHERE salary = (
        SELECT MAX(salary) 
        FROM employees
    ) 
    ORDER BY first_name, last_name
```

This syntax is equivalent to the following:

```
WITH temp_table AS (SELECT MAX(salary) FROM employees)
    SELECT employee_id, first_name, last_name, salary 
    ORDER BY first_name, last_name

```


### Joins

Joins work the same way as they did in the tidyverse. There are a few types of joins:

- "Inner joins" produce results only where the linking variable appears on both tables
- "Outer joins" produce results with all data, and fills in missing values as necessary
  + Left join keeps only observations from the first specified data if their is no match.
  + Right join keeps only observations from the second data
  + Full join keeps everything

Here's an example SQL query using two tables (movies, users)

```
SELECT users.*, count(movie_id) 
    FROM users 
    LEFT JOIN movies 
        ON users.user_id = movies.viewer_id 
    GROUP BY users.user_id 
    LIMIT 10
```

This will select the first 10 users from the users table (users.*) and add a new column with the count of the number of movies they saw.

If both tables had a field called user_id that we wanted to join on, we could say `LEFT JOIN movies ON user_id` instead of specifying
what the column is called in each table. If the column name is ambiguous (because it's present in both tables), a GROUP BY or ORDER BY or SELECT
should specify which table to take the column from by saying `table.column`

## The `sqldf` package

The main package we'll be using to convert SQL commands to R manipulations is `sqldf`.

Here's an example using the same data from the Tidy chapter.

Data comes from fivethirtyeight.com
```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(tidyverse)
mayors<-read_csv("../Datasets/Mayors.csv")
tweets<-read_csv("C:/Users/maria/OneDrive/WUSTL/2019-20/PDS/Tweets.csv")
```

`mayors` and `tweets` are tibbles, but we want to operate our SQL queries on data frames

```{r}
mayorsDF <- as.data.frame(mayors)
tweetsDF <- as.data.frame(tweets)
```

Now we can start using our newfound SQL powers to truly explore our data. For a warm-up, let's count how many mayors are present in our dataset.

```{r}
library(sqldf)
sqldf( "SELECT COUNT(DISTINCT FullName) 
            FROM mayorsDF" )
```

We can confirm this with our tidy skills:

```{r}
mayors %>% 
    select("FullName") %>%
    filter(!is.na(FullName)) %>% 
    n_distinct()
```

Note that we need to filter out NA/null values from the dataset before taking unique values in tidy. Tidy counts null as a unique value.

How about adding a column to the mayors dataset that counts the number of tweets they have made?

With SQL:
```{r}
mayors_newDF <- sqldf( "SELECT mayorsDF.*, COUNT(TweetID) AS nTweets
                            FROM mayorsDF 
                            INNER JOIN tweetsDF 
                            ON mayorsDF.TwitterHandle = tweetsDF.ScreenName 
                            GROUP BY MayorID
                            ORDER BY nTweets DESC, FullName ASC" )
head(mayors_newDF[c("MayorID", "FullName", "TwitterHandle", "nTweets")], n=10)
```

With tidy:
```{r}
tweetCounts <- tweets %>% 
    filter(!is.na(ScreenName)) %>% 
    group_by(ScreenName) %>% 
    tally()
mayors_new <- mayors %>% 
    inner_join(tweetCounts, by=c("TwitterHandle"="ScreenName")) %>% 
    group_by(MayorID) %>% 
    arrange(FullName) %>%
    arrange(desc(n)) %>% 
    select("MayorID", "FullName", "TwitterHandle", "n")
mayors_new
```