---
title: ""
output: html_document
---
## How to add code

To add code 'inline', use three ``` (back quote keys) [Don't confuse this with the apostrophe key]. After adding the three back quote keys, you will then include the language which you want to run in brackets {r}. There are various options you can use about whether you want the code to actually be displayed or evaluated. 
 
```{r setup, include =F}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
# You can set up default seting using r setup
# Here by default we used `echo = TRUE` to tell all chunks to print out the source code
# We also used eval = TRUE so that all the code is evaluated and the results displayed
# Both of these options can be overriden in a particular chunk, though
```

```{r,warning=F, fig.height=2.5, fig.width=2.5}
# You can show the code and the output and shut off warnings
a <- rnorm(100)
plot(a) 
```

---

```{r cars, eval=F}
library(fivethirtyeight)
trump_twitter$text[1:5]
# You can load in packages 
# You can choose to show the code but not evaluate it 
```
---
You can chose to evaluate a code chunk but choose not to show the code

```{r, echo=F, fig.height=2.5, fig.width=2.5}
a <- rnorm(100)
plot(a) 
```