---
title: "Lecture 3: Measures of spread and central tendency"
author: "Jacob M. Montgomery"
date: "Quantitative Political Methodology"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



----

# Lecture 3

----


## Class business

- Make sure you review online materials before lab
- Problem set 1 is due on Wed.


### Facebook and survey

- Sign up for our Facebook group: https://www.facebook.com/groups/1071702902960687/
- Take the class survey!  Can't assign teams until you all do.

https://wustl.az1.qualtrics.com/jfe/form/SV_6rpSYD3xxmbRe5v

----

## Roadmap 

Last time: 

- Populations vs. samples
- Why sample?
- How to sample
- Problems in surveys

This time: 

- Visualizing data
- Measures of central tendency and spread

----

### Pie charts = mostly aweful


```{r, echo=FALSE, out.width="1\\textwidth", fig.align='center'}
knitr::include_graphics("LectureGraphics/klass1", dpi = NA)
```

(Klass 2012)

----

### Bar charts -- more useful

```{r, echo=FALSE, out.width="1\\textwidth", fig.align='center'}
knitr::include_graphics("LectureGraphics/klass2", dpi = NA)
```

---
#### Histograms -- usually good

```{r, echo=FALSE, out.width=".75\\textwidth", fig.align='center'}
knitr::include_graphics("LectureGraphics/adams", dpi = NA)
```

(Adams 1997)


----

#### Making a histogram


\begin{center}
\begin{tabular}{rlr}
  \hline
 & State & Median Income \\ 
  \hline
1 & Alabama & 40489 \\ 
  2 & Alaska & 66953 \\ 
  3 & Arizona & 48745 \\ 
  4 & Arkansas & 37823 \\ 
  5 & California & 58931 \\ 
  6 & Colorado & 55430 \\ 
  7 & Connecticut & 67034 \\ 
  8 & Delaware & 56860 \\ 
    & \ldots & \\
  50 & Wisconsin & 49993 \\ 
  51 & Wyoming & 52664 \\ 
  52 & Puerto Rico & 18314 \\ 
   \hline
\end{tabular}
\end{center}

----

### Making a histogram

\begin{center}
\begin{tabular}{lc}
	\hline
	Class & Count \\ \hline
	\$0-\$9,999 & 0 \\
	\$10,000-\$19,999 & 1 \\
	\$20,000-\$29,999 & 0 \\
	\$30,000-\$39,999 & 3 \\
	\$40,000-\$49,999 & 27 \\
	\$50,000-\$59,999 & 14 \\
	\$60,000-\$69,999 & 7 \\
	\$70,000-\$79,999 & 0 \\ \hline
\end{tabular}
\end{center}

----

### Binning can be evil

\begin{center}
\includegraphics[width=7.5cm]{./LectureGraphics/schultz1}\\
\includegraphics[width=7.5cm]{./LectureGraphics/schultz3}\\
\footnotesize{(Ken Schultz)}\end{center}

----

### Line plots are often useful

```{r, echo=FALSE, out.width=".9\\textwidth", fig.align='center'}
knitr::include_graphics("LectureGraphics/temps", dpi = NA)
```

(NASA 2011)


----

### Can also be misleading

```{r, echo=FALSE, out.width="1\\textwidth", fig.align='center'}
knitr::include_graphics("LectureGraphics/bushgraph", dpi = NA)
```

(Klass 2012)


----


### Can also be misleading

\begin{center}
\includegraphics[height=4cm]{./LectureGraphics/bushcorr1}\includegraphics[height=4cm]{./LectureGraphics/bushcorr2}\\
\footnotesize{(Nyhan 2005)}\end{center}


----

## Looking at distributions of data

- General impressions (pattern, deviations)
- ``Shape'' (symmetric, skewed, ``bell-shaped,'' bimodal, etc.)
- ``Center'' (``midpoint'')
- ``Spread'' (range)
- ``Outliers'' (``outside overall pattern'')


----

### Our first statistics -- the mighty mean


\begin{tabular}{lrl}
& $n=$ &Sample size  \\
& $y_1, y_2, \ldots, y_n = $ &Observations\\
\end{tabular} \pause


$$\bar{y} = \displaystyle\frac{y_1 + y_2 + \ldots + y_n}{n} = \frac{1}{n}\displaystyle\sum_{i=1}^ny_i \pause \equiv \frac{1}{n}\displaystyle\sum y_i$$

----

### Test tip

\begin{tabular}{cc}
Grand mean of two samples & $\bar{y} =  \frac{n_1\bar{y}_1 +
  n_2\bar{y_2}}{n_1+n_2} $\\
\\ \pause
What if I had j samples? & $\bar{y} = \frac{n_1 \bar{y}_1 +
  n_2\bar{y}_2+\ldots+n_j \bar{y}_j}{n_1+n_2 +\ldots + n_j}$\\
\end{tabular}

----

### Finding the median of grouped data


```{r, echo=FALSE, out.width=".8\\textwidth", fig.align='center'}
knitr::include_graphics("LectureGraphics/Stats", dpi = NA)
```

----

### Test tip: Find the median of grouped data

\begin{center}
\begin{tabular}{lrrr}

\hline
Category & Frequency & Percentage & Cumulative \% \\
 \hline
  Agree Strongly &   6 & 7.4& 7.69\\ 
  Agree &   27 & 33.3& 40.7\\ 
Neither agree nor disagree &  18 & 22.2& 63.0\\ 
 Disagree &  19 & 23.5& 86.4\\ 
 Disagree Strongly &   11 & 13.6& 100.00\\ 
   \hline
\end{tabular}
\end{center}


----

### Test tips: Standard deviation ($\sqrt{Var}$)

Standard deviation:
 $$S=\sqrt{S^2} = \ \sqrt{\frac{\sum_{i=1}^n(y_i-\bar{y})^2}{n-1}}$$

\pause

- One standard deviation will be equal to something in original units (e.g., 2 inches) \pause
- $S\ge 0$ Why? \pause
- $S=0$ only if $y$ is a constant. 

----

#### Test tip on calculating variance

\begin{center}
\begin{tabular}{lll}
$y_i$ & $(y_i - \bar{y})$ & $ (y_i - \bar{y})^2$ \\
\hline
0 & -5 & 25 \\
4 & -1 & 1 \\
4 & -1 & 1 \\
5 & 0 & 0 \\
7 & 2 & 4 \\
10 & 5 & 25 \\
$\sum y_i = 30$ &$\sum(y_i-\bar{y}) = 0$&$\sum(y_i-\bar{y})^2 = 56$ \\
$\bar{y} = 5$ \\
\hline
\end{tabular}
\end{center}


- $S^2= \frac{56}{5}=11.2$
- $S=\sqrt{11.2}=3.3$




----

### A little visulaization


```{r, echo=FALSE, out.width=".8\\textwidth", fig.align='center'}
knitr::include_graphics("LectureGraphics/mmm", dpi = NA)
```

----

### Quantiles

- (55, 84, 65, 54, 61, 67, 80, 59, 81, 82)
- 54 55 59 61 65 67 80 81 82 84 \pause
- Median = $\frac{67-65}{2}$
- 50\% of the data is at \textit{or above}, and (100-50)\% is at \textit{or below}

----

#### Quantiles in R

It turns out that calculate percentiles is a more nuanced topic than I realized.  

- R includes nine different methods for calculating percentiles \pause
- We want to use 

```{r mysize=TRUE, size="\\footnotesize", tidy=FALSE, out.width=".7\\textwidth", out.height=".3\\textwidth", fig.align="center", echo=TRUE}
quantile(c(55, 84, 65, 54, 61, 67, 80, 59, 81, 82), 
         probs=c(.25, .75), 
         type=2)
```
\pause
- For more information: 
tolstoy.newcastle.edu.au/R/e17/help/att-1067/Quartiles_in_R.pdf




----

## Looking forward


```{r, echo=FALSE, out.width=".8\\textwidth", fig.align='center'}
knitr::include_graphics("LectureGraphics/population", dpi = NA)
```

