---
title: "Lecture 5: Sampling Distributions"
author: "Jacob M. Montgomery"
date: "Quantitative Political Methodology"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```





# Lecture 5

----


## Class business

- Problem set 2 will be distributed today via the syllabus
- Groups will be assigned this week.

--------

## Roadmap 

Last time: 

- Understand core concepts of probability
- Understand concept of a "parameter"
- Introduce some probability distributions


This time: 

- Understanding the concept of a sampling distribution
- Understand the concept of a standard error
- See how the CLT allows us to know the distribution of certain sample statistics.

----

## Sampling Distributions


> A **sampling distribution** is the distribution of a **statistic** given repeated sampling.  

\pause We use probability theory to derive a **distribution for a statistic**\pause, which allows us (eventually) to make inferences about **population  parameters**.  
  
  
----

### Central limit theorem

For random sampling with a **large** sample size $n$, the sampling distribution of the sample mean $\bar{y}$ is approximately normal, where $\bar{y} \sim N\left(\mu, \frac{\sigma}{\sqrt{n}}\right)$.

\pause 



- $\sigma/\sqrt{n}$ is called the standard error. \pause
    - It is the standard deviation of the sampling distribution.
    - Note that the formula includes the population standard deviation $\sigma$.
    - Pay attention or you will get them mixed up. \pause
- As $n \rightarrow \infty$, the standard error is going to get smaller and smaller.\pause
- **This** is why the normal distribution is so very important.   \pause
- Usually n=30 is ``good enough'', but it will depend on the distribution. \pause


----

### It works for EVERYTHING


```{r, echo=FALSE, out.width=".9\\textwidth", fig.align='center'}
knitr::include_graphics("LectureGraphics/CLT1", dpi = NA)
```


----

### Let's do our own experiment in R

\footnotesize
```{r mysize=TRUE, size="\\tiny", tidy=FALSE, out.width="1\\textwidth", out.height=".4\\textheight", fig.align="center", echo=TRUE}

x<-runif(100000,min = -1, max=1)
## Crazy transformation of the data
x<-sqrt(1+x)+2*x^3-23*x+abs(log(abs(x)))+2*(x>.5)+-2*(x< -.5)
## Adding a big whole in the middle with a point mass at zero
x[x>-5&x<5]<-0
plot(density(x, bw = .5), main="A crazy distribution")
```

----

\footnotesize
```{r mysize=TRUE, size="\\tiny", tidy=FALSE, out.width="1\\textwidth", out.height=".7\\textheight", fig.align="center", echo=TRUE}

# We are going to take 500 random samples 
n.samples<-1000
# The sample size is 20
sample.size<-20
# We are going to use something called a "for loop"
# First we make an empty vector to store all of our 
# sample statistics

# Create a vector filled with NA (missing data) 
# vector is of length 200
x.bars<-rep(NA, n.samples) 
# Now we are going to "loop" over the vector 1, 2, ..., 200
# in each iteration the variable "i" will increment up on value
for(i in 1:n.samples){
  # Draw a random sample
  this.sample<-sample(x, size = sample.size, replace=F) 
  # Calculate the mean and add it to the vector
  x.bars[i]<-mean(this.sample)
}
```


----

```{r mysize=TRUE, size="\\tiny", tidy=FALSE, out.width=".7\\textwidth", out.height=".4\\textwidth", fig.align="center", echo=TRUE}
plot(density(x.bars), 
     main="Sampling distribution of x.bar with n=20")
```

----

### Now with a larger sample size

\footnotesize
```{r mysize=TRUE, size="\\tiny", tidy=FALSE, out.width="1\\textwidth", out.height=".7\\textheight", fig.align="center", echo=TRUE}

# We are going to take 500 random samples 
n.samples<-1000
# The sample size is 2000
sample.size<-2000
# We are going to use something called a "for loop"
# First we make an empty vector to store all of our 
# sample statistics

# Create a vector filled with NA (missing data) 
# vector is of length 200
x.bars<-rep(NA, n.samples) 
# Now we are going to "loop" over the vector 1, 2, ..., 200
# in each iteration the variable "i" will increment up on value
for(i in 1:n.samples){
  # Draw a random sample
  this.sample<-sample(x, size = sample.size, replace=F) 
  # Calculate the mean and add it to the vector
  x.bars[i]<-mean(this.sample)
}
```

-----


```{r mysize=TRUE, size="\\tiny", tidy=FALSE, out.width=".7\\textwidth", out.height=".4\\textwidth", fig.align="center", echo=TRUE}
plot(density(x.bars), 
     main="Sampling distribution of x.bar with n=20")
```

----

### Sampling Distribution of $\bar{y}$

A key common statistic is $\bar{y} = \displaystyle \frac{1}{n} \sum y_i$ for a single sample, where $n$ is the sample size.  How is this statistic distributed?   \pause  The mean of the distribution is known to be $\mu$ (the population mean). \pause What about the spread?


----

### Standard error

The standard deviation of the sampling distribution of $\bar{y}$,  denoted $\sigma_{\bar{y}}$, is called the standard error of $\bar{y}$, and is equal to $\displaystyle \frac{\sigma}{\sqrt{n}}$.

\pause

**Under certain circumstances** we can safely assume that $\bar{y} \sim N(\mu, \frac{\sigma}{\sqrt{n}})$.

----


### Class business




- Next class we are going to learn how to calculate probabilities for a known distribution
- Then we pull it all together to make our first true inference



