---
title: "Regression Assumptions"
author: "Jacob M. Montgomery"
date: "Quantitative Political Methodology"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```





# Regression Diagnostics

## Poster questions?


-----

## Road map

Where we have been:

- Regression
- Interpreting regression
- Causal inference in regression
- Model Fit

\pause

Today:

- Regression assumptions
- How they can be broken
- Simple tricks to solve them


----


## Class Business

- PS Due today.  One more to go.
- Poster files will be due before the final lecture period.
- Optional poster session on Monday, December 11
- 1/% extra credit for the best poster as chosen by faculty
- Cookies
- Be pround of your work


-----

### Assumptions of regression

1. Observations are independent \pause
2. For any given value of x, the response y varies around the regression line according to a normal distribution. \pause
3. There is a linear relationship between x and y \pause
4. The standard deviation of y ($\sigma$) is the same for all values of x

\pause

For each one:

- What can go wrong?
- What can we do about it?


----



### Assumption 1: Observations are independent

Autocorrelation

- Time-series
- Repeated observations
- Space


\pause

Conditional standard deviation
$$\hat{\sigma} = \sqrt{\frac{SSE}{n-2}} = \sqrt{\frac{\sum (Y_i - \hat{Y_i})^2}{n-2}}$$


\pause

Standard error for $\beta$
$$\hat{\sigma}_{\hat{\beta}}= \frac{\hat{\sigma}}{\sqrt{\sum (X_i -
    \bar{X}})^2}$$ 


Your standard errors will be too small!  


----

\begin{center}
\includegraphics[scale=.8]{./LectureGraphics/GDPyear1}
\end{center}


-----


### GDP $\sim$ Year + Unemployment

\begin{center}
\includegraphics[scale=.8]{./LectureGraphics/GDPyear2}
\end{center}


-----


### GDP $\sim$ Year + Unemployment

\texttt{acf(LM1\$residuals)}

\begin{center}
\includegraphics[scale=.8]{./LectureGraphics/GDPyear3}
\end{center}

----


### Autocorrelation: Solution

- Lagged dependent variable
- ``Differencing'' dependent and/or independent variables
- Some combination thereof
- Also ... fixed effects


----


#### $\Delta$GDP $\sim$ Year + Unemployment + Lagged.GDP

\begin{center}
\includegraphics[scale=.9]{./LectureGraphics/GDPyear4}
\end{center}


----

#### Assumption 2: Response varies around line according to a normal

What can go wrong?

- Outliers
- Exacerbated by high leverage points


----

### Influential observations

\begin{center}
\includegraphics[width=100mm]{./LectureGraphics/fig58}
\end{center}


-----

#### Leverage (for bivariate regression)
$$h_i = \frac{1}{n} + \frac{(x_i - \bar{x})^2}{\sum(x - \bar{x})} $$

\pause

#### Cooks distance -- a measure of influence
$$D_i = \frac{e_i^2}{p~MSE} \left( \frac{h_i}{(1-h_i)^2}  \right)$$


\pause

- $e_i = \hat{y}_i$ - $\hat{y}_{(i)}$
- $p$ is the number variables in the model
- MSE is the mean square error of the regression


----

### Easy solutions


- Transforming variables (Xs and Ys) \pause
- Remove high-leverage outliers \pause
- Trim the variable

----

### Visualizing different Xs

\begin{center}
\includegraphics[width=4.5in]{./LectureGraphics/LogTrans1}
\end{center}

----


### Visualizing different Xs

\begin{center}
\includegraphics[width=4.5in]{./LectureGraphics/LogTrans2}
\end{center}


----


### Transforming Y's


\begin{center}
\includegraphics[width=4.5in]{./LectureGraphics/SkewedOutcome1}
\end{center}



----


### Transforming Y's


\begin{center}
\includegraphics[width=4.5in]{./LectureGraphics/SkewedOutcome2}
\end{center}

----



### Example: GDP versus per capital GDP in 1990

\begin{center}
\includegraphics[width=4.25in]{./LectureGraphics/PopGDP1}
\end{center}

----


### Example: GDP versus per capital GDP in 1990

\begin{center}
\includegraphics[width=4.25in]{./LectureGraphics/PopGDP2}
\end{center}


----

### Example: GDP versus per capital GDP in 1990

\begin{center}
\includegraphics[width=4.24in]{./LectureGraphics/PopGDP3}
\end{center}

----

### Example: GDP versus per capital GDP in 1990

\begin{center}
\includegraphics[width=4.25in]{./LectureGraphics/PopGDP4}
\end{center}


----

### Heteroscedasticity: Variance changes based on value of x


\begin{center}
\includegraphics[width=4.5in]{./LectureGraphics/Heteroscedasticity}
\end{center}

----

### Problem 4: Nonlinear relationships

\begin{center}
\includegraphics[width=4.5in]{./LectureGraphics/Nonlinear1}
\end{center}

----



### Solution: Model the relationship

\begin{center}
\includegraphics[width=4.5in]{./LectureGraphics/Nonlinear2}
\end{center}

----


### Easy regression diagnostics in R

```{r eval=FALSE, echo=TRUE}
# The first regression
FL.LM1 <- lm(gdpen~pop, data=FL)
plot(FL.LM1)


# The second regression
FL.LM2 <- lm(log(gdpen)~log(pop), data=FL)
plot(FL.LM2)
```


----


### Bad models

\begin{center}
\includegraphics[width=3.1in]{./LectureGraphics/ResidPlot1}
\end{center}



----


### Bad models

\begin{center}
\includegraphics[width=3.1in]{./LectureGraphics/ResidPlot2}
\end{center}


----

### Bad models

\begin{center}
\includegraphics[width=3.1in]{./LectureGraphics/ResidPlot3}
\end{center}


----

### Bad models

\begin{center}
\includegraphics[width=3.1in]{./LectureGraphics/ResidPlot4}
\end{center}




----

### Better Model

\begin{center}
\includegraphics[width=3.1in]{./LectureGraphics/ResPlot5}
\end{center}


----

### Better Model

\begin{center}
\includegraphics[width=3.1in]{./LectureGraphics/ResPlot6}
\end{center}


----

### Better Model

\begin{center}
\includegraphics[width=3.1in]{./LectureGraphics/ResPlot7}
\end{center}


----

### Better Model

\begin{center}
\includegraphics[width=3.1in]{./LectureGraphics/ResPlot8}
\end{center}

